---
title: "Human-driven trophic simplification in Azorean lake ecosystems"
output:
  pdf_document: default
---

**About this script.** This R Markdown reproduces the analyses for the manuscript on trophic structure change in Azorean lakes over the past ~2000 years. It loads data (DCA scores, guild abundances, climate reconstructions, pollen), runs variance partitioning (phase-based and moving window), sensitivity analyses (phase boundaries, leave-one-out, null models), and produces main and supplementary figures. Custom helper functions are in `functions/source_custom_functions.R`. Run chunks in order; all packages are loaded in the first code section.

```{r}
# Set-up chunk
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.width = 12
)
```

# Load libraries
```{r sources, include=FALSE}
# ------------------------------------------------------------------------------
# All dependencies are loaded once here. Do not add library() in later chunks;
# if a package is needed, add it below and re-run this chunk.
# ------------------------------------------------------------------------------

# -------------------------------
# 🧰 General-Purpose & Data Handling
# -------------------------------
library(tidyverse)      # Loads: ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
library(zoo)            # Time-series objects and interpolation
library(parallel)       # Base R parallel processing support
library(furrr)          # Parallel purrr (requires future backend)
library(sf)             # Spatial vector data handling

# -------------------------------
# 🖍 Aesthetic & Utility Enhancements
# -------------------------------
library(scales)         # Useful for custom ggplot2 axes
library(png)            # For reading PNG images (only if you're embedding raster images)
library(ggtext)         # Rich text/markdown/emoji in ggplot2 elements
library(Hmisc)          # For `capitalize()` or advanced summary stats
library(scico)          # Color palettes (colorblind-safe, perceptually uniform)
library(conflicted)     # Safer namespace conflict management

# -------------------------------
# 📊 Visualization & Plot Layout
# -------------------------------
# ggplot2 already loaded via tidyverse, but other extensions aren't:
library(ggrepel)        # Smart label positioning
library(ggpmisc)        # Annotations and statistical labels
library(viridis)        # Better color palettes
library(cowplot)        # Clean multi-panel figures
library(patchwork)      # Combine ggplots with intuitive syntax
library(gridExtra)      # Arrange base or ggplots in grids
library(grid)           # Base graphics system (already available, no need to load)
library(ggnewscale)     # Multiple color/fill scales
library(ggridges)       # Ridgeline plots
library(ggh4x)          # Extended faceting and axes
library(ggspatial)      # North arrows, scale bars, etc.
library(ggordiplots)    # Ordination diagrams (RDA, CCA, NMDS)

# -------------------------------
# 📈 Statistical Analysis & Modeling
# -------------------------------
library(vegan)          # Ecological ordination (NMDS, CCA, RDA), diversity metrics
library(codyn)          # Temporal beta-diversity and dynamics
library(broom)          # Tidy model outputs
library(mgcv)           # Generalized Additive Models
library(scam)           # Shape-constrained additive models
library(gratia)         # Diagnostics and visualization for `mgcv::gam()`

# -------------------------------
# 🤖 Machine Learning & Interpretation
# -------------------------------
library(caret)          # Framework for ML model training/testing
library(e1071)          # SVMs, Naive Bayes, etc. (caret dependency)
library(evtree)         # Evolutionary trees
library(rpart)          # Classification/regression trees
library(randomForest)   # Random Forests
library(rpart.plot)     # Visualize rpart trees
library(earth)          # MARS regression
library(DALEX)          # Model explanation tools
library(DALEXtra)       # Extra methods for xgboost, caret, etc.
library(irr)            # Inter-rater agreement (Cohen's Kappa, etc.)
library(ape)            # Phylogenetic tools (tree manipulation, distances)

# -------------------------------
# 📂 Custom Functions
# -------------------------------
source("functions/source_custom_functions.R")  # Custom paleoecological tools

# -------------------------------
# 🚨 Solve Function Conflicts
# -------------------------------
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("summarise", "dplyr")
conflict_prefer("summarize", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("explain", "DALEX")
conflict_prefer("annotate", "ggplot2")

# -------------------------------
# Output folders (create if missing)
# -------------------------------
dir.create("figures", showWarnings = FALSE, recursive = TRUE)
dir.create("supplementary", showWarnings = FALSE, recursive = TRUE)
dir.create("outputs", showWarnings = FALSE, recursive = TRUE)
```

# Load data
```{r}

load("data/clean_source_data_files.RData")

# Define functional groups (used throughout trophic structure analyses)
alternative_functional_groups_guilds_globi_merged <- c(
  "high_profile", "low_profile", "motile", "euplanktonic",
  "algivore", "detritivore", "plantivore", "predator"
)

# ------------------------------------------------------------------------------
# Load lake metadata
# ------------------------------------------------------------------------------
lake_metadata <- read.csv(file = "data/table_lake_metadata.csv")  # lake coordinates, area, etc.

# ------------------------------------------------------------------------------
# Load guild relative abundances and diversity tables
# ------------------------------------------------------------------------------
norm_abund_guilds <- read.table("data/table_abundance_guilds.csv",header = TRUE, sep = ",", dec = ".")
div_guilds <- read.table("data/table_diversity_guilds.csv", header = TRUE, sep = ",", dec = ".")

# ------------------------------------------------------------------------------
# Load climate and environmental reconstructions
# ------------------------------------------------------------------------------
nao_hernandez <- read.table("data/table_nao_hernandez.csv", header = TRUE, sep = ",", dec = ".")
nhst_buntgen <- read.table("data/table_nhst_buntgen.csv", header = TRUE, sep = ",", dec = ".")

# ------------------------------------------------------------------------------
# Load pollen group data
# ------------------------------------------------------------------------------
pollen <- read.table("data/table_pollen_azores.csv", header = TRUE, sep = ",", dec = ".")

```

# Analysis - Detrended Correspondence Analysis
## DCA Diatoms

```{r fig.asp = 1, fig.width = 12}
# # ------------------------------------------------------------------------------
# # Extract PCoA vectors (1st and 2nd axes) for each lake
# # ------------------------------------------------------------------------------
# 
# ls_df_pcoa_scores_diatoms_hellinger <- ls_df_diat_wide_codes %>%
#   # Remove metadata columns
#   map(., ~ .x %>% select(-c(lake:core_depth_id))) %>%
#   # Replace NAs with zeros
#   map(., ~ .x %>% replace(is.na(.), 0)) %>%
#   # Remove species with total abundance = 0 across all samples
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   # Apply Hellinger transformation
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   # Compute Bray–Curtis dissimilarity
#   map(., ~ .x %>% vegdist(method = "bray", na.rm = TRUE)) %>%
#   # Run PCoA
#   map(., ~ .x %>% pcoa()) %>%
#   # Extract eigenvectors
#   map(., function(x) x$vectors) %>%
#   # Convert first two axes into a data frame
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # ------------------------------------------------------------------------------
# # Bind metadata and PCoA scores for each lake
# # ------------------------------------------------------------------------------
# 
# ls_df_all_scores_diatoms_amd_globi <- NULL
# 
# for (i in 1:length(ls_df_pcoa_scores_diatoms_hellinger)) {
#   ls_df_all_scores_diatoms_amd_globi[[i]] <- bind_cols(
#     ls_df_codes_diat[[i]],
#     ls_df_pcoa_scores_diatoms_hellinger[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Fix decimal comma issue in depth field
# # ------------------------------------------------------------------------------
# 
# ls_df_all_scores_diatoms_amd_globi <- ls_df_all_scores_diatoms_amd_globi %>%
#   map(., ~ .x %>% mutate(dec_depth = as.numeric(str_replace(dec_depth, ",", "."))))
# 
# # ------------------------------------------------------------------------------
# # Calculate abundance, richness, and Simpson index for each sample
# # ------------------------------------------------------------------------------
# 
# ls_df_diatoms_wide_abund <- ls_df_diat_wide_codes %>%
#   # Select species-only columns
#   map(., ~ .x %>% select(8:last_col())) %>%
#   # Add total abundance column
#   map(., ~ .x %>% mutate(lake_abundance = rowSums(.x))) %>%
#   # Add richness (count of present species)
#   map(., ~ .x %>% mutate(local_richness = apply(.x[8:(ncol(.x) - 1)] > 0, 1, sum))) %>%
#   # Add Simpson diversity index
#   map(., ~ .x %>% mutate(simpson = apply(.x[8:(ncol(.x) - 1)], 1, function(x) abdiv::simpson(x)))) %>%
#   # Keep only calculated metrics
#   map(., ~ .x %>% select(lake_abundance, local_richness, simpson))
# 
# # ------------------------------------------------------------------------------
# # Bind PCoA scores with richness and abundance metrics
# # ------------------------------------------------------------------------------
# 
# for (i in 1:length(ls_df_diatoms_wide_abund)) {
#   ls_df_all_scores_diatoms_amd_globi[[i]] <- bind_cols(
#     ls_df_all_scores_diatoms_amd_globi[[i]],
#     ls_df_diatoms_wide_abund[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Extract DCA vectors (1st and 2nd axes) for each lake
# # ------------------------------------------------------------------------------
# 
# ls_dca_diatoms <- ls_df_diat_wide_codes %>%
#   # Remove metadata columns
#   map(., ~ .x %>% select(-c(lake:core_depth_id))) %>%
#   # Remove species with zero abundance across all samples
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   # Apply Hellinger transformation
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   # Perform DCA
#   map(., ~ .x %>% decorana()) %>%
#   # Extract DCA scores
#   map(., function(x) scores(x)) %>%
#   # Convert first two axes into a data frame
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # ------------------------------------------------------------------------------
# # Bind DCA scores to existing PCoA + richness/abundance data
# # ------------------------------------------------------------------------------
# 
# for (i in 1:length(ls_df_all_scores_diatoms_amd_globi)) {
#   ls_df_all_scores_diatoms_amd_globi[[i]] <- bind_cols(
#     ls_df_all_scores_diatoms_amd_globi[[i]],
#     ls_dca_diatoms[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Create additional list with DCA scores + metadata
# # ------------------------------------------------------------------------------
# 
# ls_dca_diatoms_2 <- ls_df_diat_wide_codes %>%
#   map(., ~ .x %>% select(-c(lake:core_depth_id))) %>%
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   map(., ~ .x %>% decorana()) %>%
#   map(., function(x) scores(x)) %>%
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# for (i in 1:length(ls_dca_diatoms_2)) {
#   ls_dca_diatoms_2[[i]] <- bind_cols(
#     ls_df_codes_diat[[i]],
#     ls_dca_diatoms_2[[i]]
#   )
# }

```

## DCA Chironomids
```{r fig.asp = 1, fig.width = 12}

# # ------------------------------------------------------------------------------
# # Extract PCoA vectors (1st and 2nd axes) for each lake based on chironomid data
# # ------------------------------------------------------------------------------
# 
# ls_df_pcoa_scores_chiro_hellinger <- ls_df_chiro_wide_codes %>%
#   # Remove metadata columns
#   map(., ~ .x %>% select(-c(lake:depth_top))) %>%
#   # Remove species columns with no occurrences
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   # Apply Hellinger transformation
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   # Compute Bray–Curtis dissimilarity matrix
#   map(., ~ .x %>% vegdist(method = "bray", na.rm = TRUE)) %>%
#   # Perform PCoA
#   map(., ~ .x %>% pcoa()) %>%
#   # Extract eigenvectors
#   map(., function(x) x$vectors) %>%
#   # Keep first two axes and convert to data frame
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # ------------------------------------------------------------------------------
# # Combine metadata with PCoA scores
# # ------------------------------------------------------------------------------
# 
# ls_df_all_scores_chiro_amd_globi <- NULL
# 
# for (i in 1:length(ls_df_pcoa_scores_chiro_hellinger)) {
#   ls_df_all_scores_chiro_amd_globi[[i]] <- bind_cols(
#     df_chiro_codes[[i]],
#     ls_df_pcoa_scores_chiro_hellinger[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Assign lake names to list entries
# # ------------------------------------------------------------------------------
# 
# names(ls_df_all_scores_chiro_amd_globi) <- unlist(get_names_list(ls_df_all_scores_chiro_amd_globi))
# 
# # ------------------------------------------------------------------------------
# # Calculate chironomid abundance, richness, and Simpson index
# # ------------------------------------------------------------------------------
# 
# lake_abund_chiro <- ls_df_chiro_wide_codes %>%
#   # Keep only species columns
#   map(., ~ .x %>% select(8:last_col())) %>%
#   # Compute total abundance per sample
#   map(., ~ .x %>% mutate(chiro_lake_abundance = rowSums(.x))) %>%
#   # Compute local richness (number of present species)
#   map(., ~ .x %>% mutate(chiro_local_richness = apply(.x[8:(ncol(.x) - 1)] > 0, 1, sum))) %>%
#   # Compute Simpson diversity index
#   map(., ~ .x %>% mutate(chiro_simpson = apply(.x[8:(ncol(.x) - 1)], 1, function(x) abdiv::simpson(x)))) %>%
#   # Retain calculated metrics only
#   map(., ~ .x %>% select(chiro_lake_abundance, chiro_local_richness, chiro_simpson))
# 
# # ------------------------------------------------------------------------------
# # Bind abundance and diversity metrics to main list
# # ------------------------------------------------------------------------------
# 
# for (i in 1:length(ls_df_all_scores_chiro_amd_globi)) {
#   ls_df_all_scores_chiro_amd_globi[[i]] <- bind_cols(
#     ls_df_all_scores_chiro_amd_globi[[i]],
#     lake_abund_chiro[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Extract DCA vectors (1st and 2nd axes) from chironomid data
# # ------------------------------------------------------------------------------
# 
# ls_dca_chiro <- ls_df_chiro_wide_codes %>%
#   # Remove metadata columns
#   map(., ~ .x %>% select(-c(lake:depth_top))) %>%
#   # Remove species columns with no data
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   # Apply Hellinger transformation
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   # Perform DCA
#   map(., ~ .x %>% decorana()) %>%
#   # Extract site scores
#   map(., function(x) scores(x)) %>%
#   # Keep first two axes and convert to data frame
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # ------------------------------------------------------------------------------
# # Bind DCA scores to metadata + PCoA + diversity info
# # ------------------------------------------------------------------------------
# 
# for (i in 1:length(ls_df_all_scores_chiro_amd_globi)) {
#   ls_df_all_scores_chiro_amd_globi[[i]] <- bind_cols(
#     ls_df_all_scores_chiro_amd_globi[[i]],
#     ls_dca_chiro[[i]]
#   )
# }
# 
# # ------------------------------------------------------------------------------
# # Create second DCA list with metadata attached
# # ------------------------------------------------------------------------------
# 
# ls_dca_chiro_2 <- ls_df_chiro_wide_codes %>%
#   map(., ~ .x %>% select(-c(lake:core_depth_id))) %>%
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   map(., ~ .x %>% decorana()) %>%
#   map(., function(x) scores(x)) %>%
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# for (i in 1:length(ls_dca_chiro_2)) {
#   ls_dca_chiro_2[[i]] <- bind_cols(
#     df_chiro_codes[[i]],
#     ls_dca_chiro_2[[i]]
#   )
# }


```

## DCA Guilds
```{r fig.asp = 1, fig.width = 12}
# # ------------------------------------------------------------------------------
# # Clean initial data and remove NA rows
# # ------------------------------------------------------------------------------
# 
# df_fgroups <- df_fgroups[!is.na(rowSums(df_fgroups[, -c(1:4)])), ]
# 
# # Ungroup any groupings that might exist in df_fgroups
# df_fg <- df_fgroups %>% ungroup()
# 
# # Extract identifying metadata (lake, core_id, depth, age_ce)
# df_fg_codes <- df_fg %>%
#  ungroup() %>%
#  select(lake:age_ce) %>%
#  distinct()
# 
# # Split the metadata by lake into a list
# ls_df_fgroups_wide_codes <- df_fg_codes %>%
#  group_split(lake)
# 
# # Assign names to list based on lake
# names(ls_df_fgroups_wide_codes) <- unlist(get_names_list(ls_df_fgroups_wide_codes))
# 
# # ------------------------------------------------------------------------------
# # Run DCA on each lake's functional group composition
# # ------------------------------------------------------------------------------
# 
# ls_dca_fgroups <- df_fg %>%
#  group_split(lake) %>%
#  # Remove metadata columns
#  map(., ~ .x %>% dplyr::select(-c(lake:age_ce))) %>%
#  # Remove empty rows (samples with no groups)
#  map(., ~ .x %>% filter(rowSums(.) != 0, na.rm = TRUE)) %>%
#  # Remove empty group columns (groups not present)
#  map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#  # Apply Hellinger transformation
#  map(., ~ .x %>% decostand(method = "hellinger")) %>%
#  # Perform DCA ordination
#  map(., ~ .x %>% decorana()) %>%
#  # Extract sample scores
#  map(., function(x) scores(x)) %>%
#  # Convert first two axes to data frame
#  map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # Also create a list of raw data split by lake (used later)
# ls_dca_fgroups_EWS <- df_fg %>% group_split(lake)
# 
# # ------------------------------------------------------------------------------
# # Combine metadata and DCA results
# # ------------------------------------------------------------------------------
# 
# # Combine all DCA outputs into one data frame and bind with metadata
# df_dca_fgroups_codes_wide <- ls_dca_fgroups %>%
#  bind_rows() %>%
#  bind_cols(df_fg_codes, .)
# 
# # Split DCA results with metadata by lake
# ls_df_dca_fgroups_codes_wide <- df_dca_fgroups_codes_wide %>%
#  group_split(lake)
# 
# # Assign names to list
# names(ls_df_dca_fgroups_codes_wide) <- unlist(get_names_list(ls_df_dca_fgroups_codes_wide))
# 
# # ------------------------------------------------------------------------------
# # Prepare input for PCoA analyses
# # ------------------------------------------------------------------------------
# 
# # Split raw abundance data by lake
# ls_df_fg <- df_fg %>%
#  group_split(lake)
# 
# # Assign names to list
# names(ls_df_fg) <- unlist(get_names_list(ls_df_fg))
# 
# # ------------------------------------------------------------------------------
# # Compute PCoA for each lake using Bray–Curtis distances
# # ------------------------------------------------------------------------------
# 
# ls_df_pcoa_scores_fgroups <- ls_df_fg %>%
#   # Remove metadata columns
#   map(., ~ .x %>% select(-c(lake:age_ce))) %>%
#   # Replace NAs with 0s
#   map(., ~ .x %>% replace(is.na(.), 0)) %>%
#   # Remove empty group columns
#   map(., ~ .x %>% select_if(list(~ sum(.x) > 0))) %>%
#   # Optional: Apply Hellinger transformation (common before Bray-Curtis, but not mandatory here)
#   map(., ~ .x %>% decostand(method = "hellinger")) %>%
#   # Calculate Bray–Curtis distance
#   map(., ~ .x %>% vegdist(method = "bray", na.rm = TRUE)) %>%
#   # Perform PCoA
#   map(., ~ .x %>% pcoa()) %>%
#   # Extract sample scores
#   map(., function(x) x$vectors) %>%
#   # Keep first two PCoA axes and convert to data frame
#   map(., function(x) as.data.frame(unlist(x[, 1:2])))
# 
# # ------------------------------------------------------------------------------
# # Bind metadata and PCoA scores
# # ------------------------------------------------------------------------------
# 
# ls_df_fg_codes <- ls_df_fg %>% map(., ~ .x %>% select(c(lake:age_ce)))
# 
# for (i in 1:length(ls_df_pcoa_scores_fgroups)) {
#   ls_df_pcoa_scores_fgroups[[i]] <- bind_cols(
#     ls_df_fg_codes[[i]],
#     ls_df_pcoa_scores_fgroups[[i]]
#   )
# }

```

## Combine DCAs
```{r fig.asp = 1, fig.width = 12}

# # --------------------------------------------
# # Apply axis correction to diatom DCA scores
# # --------------------------------------------
# b_ls_dca_diatoms <- 
#   ls_dca_diatoms_2 %>%
#   map(~ .x %>% arrange(age_ce)) %>%         # Sort by age
#   map(~ .x %>% convert_DCA()) %>%           # Flip DCA if needed (Custom function in functions file)
#   bind_rows() %>%
#   select(lake, age_ce, DCA1, DCA2)
# 
# # Fix Prata manually (function failed or returned incorrect orientation)
# b_ls_dca_diatoms[b_ls_dca_diatoms$lake == "Prata", "DCA1"] <- 
#   b_ls_dca_diatoms[b_ls_dca_diatoms$lake == "Prata", "DCA1"] * -1
# 
# # --------------------------------------------
# # Apply axis correction to chironomid DCA scores
# # --------------------------------------------
# b_ls_dca_chiro <- 
#   ls_dca_chiro_2 %>%
#   map(~ .x %>% arrange(age_ce)) %>%
#   map(~ .x %>% convert_DCA()) %>%
#   bind_rows() %>%
#   select(lake, age_ce, DCA1, DCA2)
# 
# # --------------------------------------------
# # Apply axis correction to functional group DCA scores
# # --------------------------------------------
# b_ls_dca_fgroup <- 
#   ls_df_dca_fgroups_codes_wide %>%
#   map(~ .x %>% arrange(age_ce)) %>%
#   map(~ .x %>% convert_DCA()) %>%
#   bind_rows() %>%
#   select(lake, age_ce, DCA1, DCA2)
# 
# --------------------------------------------
# Combine all DCA scores into a single long-format dataframe
# --------------------------------------------
df_dca_multi <- bind_rows(
 b_ls_dca_diatoms,
 b_ls_dca_chiro,
 b_ls_dca_fgroup,
 .id = "troph"  # 1 = diatoms, 2 = chiro, 3 = fgroups
) %>%
 mutate(group = as.factor(troph))  # Ensure group is treated as factor for plotting

# --------------------------------------------
# Split data by lake into a list for faceted plotting
# --------------------------------------------
ls_dca_multi <- df_dca_multi %>%
 group_split(lake)

# Assign lake names as list names
names(ls_dca_multi) <- unlist(get_names_list(ls_dca_multi))

# --------------------------------------------
# Create list of plots comparing diatoms vs chironomids by lake
# --------------------------------------------
ls_plot_df_dca_multi_all <- 
 ls_dca_multi %>%
 map(~ .x %>% filter(group %in% 1:2)) %>%  # Filter to diatoms (1) and chironomids (2)
 map2(
   ., names(ls_dca_multi),
   ~ ggplot(.x, aes(x = age_ce, y = DCA1)) +
     geom_smooth(
       aes(colour = group), 
       linewidth = 2, alpha = 0.5, method = "gam", show.legend = TRUE
     ) +
     # Optional: Add raw points if desired
     # geom_point(aes(color = group), size = 3) +
     scale_colour_viridis_d(
       name = "Groups",
       labels = c("Producers", "Consumers")
     ) +
     theme(plot.background = NULL) +
     labs(title = .y) +  # Lake name as plot title
     theme_bw() +
     theme(
       plot.title = element_text(face = "bold", size = 12),
       legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
       legend.justification = c(0, 1),
       legend.position.inside = c(0, 1),
       axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
       panel.grid.major = element_line(colour = "grey70", linewidth = 0.2),
       panel.grid.minor = element_blank()
     )
 )


```

# Analysis - Community Trophic Structure Analysis
## Get clusters 
```{r fig.asp = 0.5, fig.width = 12}

# --------------------------------------------
# Combine list of functional group data frames into one table
# --------------------------------------------
tab_sums <- 
  ls_df_fg %>% 
  map(~.x %>% as.data.frame()) %>%  # Convert each list element to data frame
  bind_rows()  # Combine into one unified data frame

# --------------------------------------------
# Normalize abundance data by row (i.e., relative abundance)
# --------------------------------------------
norm_abund <- 
  bind_cols(
    select(tab_sums, lake:age_ce),  # Retain identifying metadata
    select(tab_sums, high_profile:euplanktonic) / rowSums(select(tab_sums, high_profile:euplanktonic)),  # Normalize functional groups
    select(tab_sums, algivore:last_col()) / rowSums(select(tab_sums, algivore:last_col()))  # Normalize trait groups
  ) %>% 
  replace(is.na(.), 0)  # Replace any resulting NAs with 0s

# --------------------------------------------
# Detect available CPU cores for parallel processing
# --------------------------------------------
numCores <- parallel::detectCores()

# --------------------------------------------
# 4. Run clustering optimization to identify best AMD cluster number
# --------------------------------------------
ls_opt_clusts <- 
  norm_abund %>%
  select(-c(lake:age_ce)) %>%  # Remove metadata columns
  PlotoptAMDclusters(., .iterations = 1000, .min_clusts = 3, .num_groups = 9)  # Evaluate from 3 to 9 clusters

# --------------------------------------------
# Identify optimal number of clusters via asymptote detection
# --------------------------------------------
x <- ls_opt_clusts$results$.clusters
y <- ls_opt_clusts$results$.maxpmean
dy <- diff(y)  # Rate of change in model performance
threshold <- 0.01  # Define a threshold for plateau detection
asymptote_point <- which(abs(dy) < threshold)[1]  # First point where improvement stabilizes

# Visual check for asymptote
plot(x, y, type = "l", main = "Identifying Asymptote")
points(x[asymptote_point], y[asymptote_point], col = "red", pch = 19, cex = 1.5)
abline(v = x[asymptote_point], col = "blue", lty = 2)
text(x[asymptote_point], y[asymptote_point], labels = paste0("x = ", x[asymptote_point]), pos = 4)

# Extract cluster number at asymptote
ls_opt_clusts <- as.vector(ls_opt_clusts$results[asymptote_point, 1])$.clusters

# --------------------------------------------
# Assign cluster identity to each row (sample)
# --------------------------------------------
opt_AMD_clusts_fgroups_paleo <- 
  norm_abund %>%
  select(-c(lake:age_ce)) %>%  # Remove metadata
  select_if(~ sum(.) > 0) %>%  # Drop zero-only columns
  getAMDclusters(.data = ., .opt_num_clusts = ls_opt_clusts, .iterations = 5000) %>% 
  dplyr::rename(amd_clusts = .clst)  # Rename cluster column

# Ensure cluster is a proper factor
opt_AMD_clusts_fgroups_paleo <- opt_AMD_clusts_fgroups_paleo %>%
  mutate(amd_clusts = factor(as.vector(unlist(.[, "amd_clusts"]))))

# --------------------------------------------
# Prepare ordination scores and metadata
# --------------------------------------------
glob_df_pcoa_scores_fgroups <- bind_rows(ls_df_pcoa_scores_fgroups)

# Subset ordination scores to match samples in norm_abund
sub_glob_df_pcoa_scores_fgroups <- 
  glob_df_pcoa_scores_fgroups %>% 
  filter(core_depth_id %in% norm_abund$core_depth_id)

# --------------------------------------------
# Merge metadata, ordination, and cluster assignment into one table
# --------------------------------------------
global_matrix_merged_paleo <- 
  bind_cols(
    select(norm_abund, lake:age_ce),
    select(sub_glob_df_pcoa_scores_fgroups, DCA1:DCA2),
    amd_clusts = factor(opt_AMD_clusts_fgroups_paleo[, 1])  # Add cluster assignment
  ) %>% 
  left_join(., 
            select(ungroup(df_fgroups_glob_div), core_depth_id:total_nspp_by_lake_core) %>% 
              distinct(),
            by = "core_depth_id")

# --------------------------------------------
# Create color mapping key based on cluster means
# --------------------------------------------
key <- 
  bind_cols(
    global_matrix_merged_paleo,
    select(norm_abund, c(high_profile:last_col()))
  ) %>% 
  group_by(amd_clusts) %>% 
  summarise(
    mean_euplanktonic = mean(euplanktonic, na.rm = TRUE),
    mean_total_nspp_by_lake_core = mean(total_nspp_by_lake_core, na.rm = TRUE)
  ) %>% 
  arrange(mean_total_nspp_by_lake_core) %>% 
  mutate(rank = row_number())  # Assign ranking based on richness

# --------------------------------------------
# Apply ranking to cluster IDs (relabel clusters by complexity)
# --------------------------------------------
cluster_mapping <- key %>% 
  select(amd_clusts, rank) %>% 
  deframe()  # Convert to named vector

# Replace cluster IDs with rank values
global_matrix_merged_paleo <- global_matrix_merged_paleo %>%
  mutate(amd_clusts = cluster_mapping[as.character(amd_clusts)]) %>%
  mutate(amd_clusts = as.factor(amd_clusts))

# --------------------------------------------
# [Decision made here] Merge two cluster levels
# --------------------------------------------
global_matrix_merged_paleo$amd_clusts[global_matrix_merged_paleo$amd_clusts == 6] <- 5

# --------------------------------------------
# Split final data frame by lake into a named list
# --------------------------------------------
ls_all_hclusts_complete_paleo <- 
  global_matrix_merged_paleo %>%
  group_split(lake)

# Assign names to list elements using helper function
names(ls_all_hclusts_complete_paleo) <- unlist(get_names_list(ls_all_hclusts_complete_paleo))


```

## Diversity by community trophic clusters
```{r fig.asp=1, fig.width=12}

# ----------------------------------------------
# Merge all lake-specific dataframes into one long-format table
# ----------------------------------------------
merged_data <- ls_all_hclusts_complete_paleo %>%
  map(~.x %>% as.data.frame()) %>%
  bind_rows() %>%
  filter(age_ce > 0) %>%  # Remove non-positive ages (pre-zero)
  select(lake, core_depth_id, amd_clusts, total_nspp_by_lake_core) %>%
  unique()

# ----------------------------------------------
# Perform ANOVA and Tukey post-hoc test
# ----------------------------------------------
anova_res <- aov(total_nspp_by_lake_core ~ amd_clusts, data = merged_data)
tukey_res <- TukeyHSD(anova_res)

# Generate compact letter display to indicate group differences
cld <- multcompView::multcompLetters4(anova_res, tukey_res)

# Extract significant letters and prepare for plotting
signif_letters <- data.frame(
  amd_clusts = names(cld$amd_clusts$Letters),
  letters = cld$amd_clusts$Letters
)

# Add y-position for each letter above boxplots
signif_letters <- signif_letters %>%
  left_join(
    merged_data %>%
      group_by(amd_clusts) %>%
      summarise(max_y = max(total_nspp_by_lake_core, na.rm = TRUE)),
    by = "amd_clusts"
  ) %>%
  mutate(y_position = max_y * 1.1)

# ----------------------------------------------
# Create boxplot comparing diversity across clusters (all lakes combined)
# ----------------------------------------------
box_plot_clusts <- ggplot(merged_data, aes(x = amd_clusts, y = total_nspp_by_lake_core)) +
  geom_boxplot(aes(fill = amd_clusts)) +
  geom_text(
    aes(x = amd_clusts, y = y_position, label = letters),
    inherit.aes = FALSE,
    data = signif_letters, size = 6
  ) +
  ggpubr::stat_compare_means(method = "anova", label.y = 9, cex = 4) +  # ANOVA p-value label
  scale_fill_viridis_d(name = "Community Trophic \nstructures", option = "viridis", direction = -1) +
  theme_minimal() +
  labs(x = "Community Trophic Structures", y = "Number of species") +
  theme(
    plot.title = element_text(size = 12),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.justification = c(0, 1),
    legend.position = "bottom",
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Assuming your x-axis variable is called 'x_var' in fig_2_time_clusts data:
box_plot_clusts$data$amd_clusts <- factor(box_plot_clusts$data$amd_clusts, 
                                       levels = c(6,5,4,3,2,1))
box_plot_clusts_mod <- 
  box_plot_clusts + 
  scale_fill_viridis_d(direction = 1) +
  theme(legend.position = "none")

# Rebuild the plot to reflect the new x-axis order
box_plot_clusts_mod <- box_plot_clusts_mod +
  scale_x_discrete()  # Ensures x-axis reflects the new order

# Create the inset plot
box_plot_clusts_mod <- box_plot_clusts_mod +
  #theme(plot.background = element_rect(             # Add a box around the plot
  #  color = "grey", size = 0.5, fill = NA            # Black border, no fill
  #)) +
guides("none")

```

## Lake-scale proportions trophic structures 
```{r fig.asp = 1, fig.width = 12}

# ------------------------------------------------------------------------------
# Define vertical reference lines for key historical periods
# ------------------------------------------------------------------------------
highlighted_dates <- c(750, 1050, 1450, 1750)

# ------------------------------------------------------------------------------
# Combine clustering outputs from all lakes into a single data frame
# ------------------------------------------------------------------------------
data_combined <- ls_all_hclusts_complete_paleo %>%
  map_dfr(as.data.frame)

# ------------------------------------------------------------------------------
# Compute time coverage per lake and order lakes by coverage length
# ------------------------------------------------------------------------------
lake_ranges <- data_combined %>%
  group_by(lake) %>%
  summarise(
    min_age = min(age_ce, na.rm = TRUE),
    max_age = max(age_ce, na.rm = TRUE),
    data_range = max_age - min_age,
    .groups = "drop"
  ) %>%
  arrange(desc(data_range))

lake_levels <- lake_ranges$lake  # Preserve lake order for plotting

# ------------------------------------------------------------------------------
# Set lake as ordered factor and cluster as categorical
# ------------------------------------------------------------------------------
data_combined <- data_combined %>%
  mutate(
    lake = factor(lake, levels = lake_levels),
    amd_clusts = as.factor(amd_clusts)
  )

# ------------------------------------------------------------------------------
# Define shaded rectangles to mask periods without data
# ------------------------------------------------------------------------------
rect_data <- lake_ranges %>%
  select(lake, min_age, max_age) %>%
  mutate(
    lake = factor(lake, levels = lake_levels),
    xmin1 = 0,
    xmax1 = min_age,
    xmin2 = max_age,
    xmax2 = 2010
  ) %>%
  pivot_longer(c(xmin1, xmin2), names_to = "rect_part", values_to = "xmin") %>%
  pivot_longer(c(xmax1, xmax2), names_to = "rect_part2", values_to = "xmax") %>%
  filter(substr(rect_part, 5, 5) == substr(rect_part2, 5, 5)) %>%
  select(lake, xmin, xmax) %>%
  mutate(ymin = -Inf, ymax = Inf)

# ------------------------------------------------------------------------------
# Keep only lakes with ≥2 trophic clusters
# ------------------------------------------------------------------------------
lake_valid_clusters <- data_combined %>%
  distinct(lake, amd_clusts) %>%
  count(lake) %>%
  filter(n >= 2) %>%
  pull(lake)

# ------------------------------------------------------------------------------
# Final filtering: remove sparse groups and ensure valid lake set
# ------------------------------------------------------------------------------
data_combined_clean <- data_combined %>%
  filter(
    lake %in% lake_valid_clusters,
    !is.na(age_ce), !is.na(amd_clusts),
    is.finite(age_ce)
  ) %>%
  group_by(lake, amd_clusts) %>%
  filter(n() >= 3, sd(age_ce, na.rm = TRUE) > 1) %>%
  ungroup()

rect_data_clean <- rect_data %>%
  filter(lake %in% lake_valid_clusters)

# ------------------------------------------------------------------------------
# Build final plot
# ------------------------------------------------------------------------------
plot_density_fclusts_time_lines_1B <- 
  ggplot(data_combined_clean, aes(x = age_ce, fill = amd_clusts)) +
  facet_wrap(~ lake, ncol = 1, scales = "fixed", strip.position = "right") +
  geom_density(adjust = 1.5, position = "fill", colour = "white") +
  geom_rect(
    data = rect_data_clean,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "white", colour = "white"
  ) +
  geom_vline(xintercept = highlighted_dates, colour = "white", linetype = "dashed") +
  scale_fill_viridis_d(
    name = "Community Trophic \nstructures", option = "viridis", direction = -1
  ) +
  lims(x = c(0, 2010)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), labels = c("0", "0.5", "1")) +
  theme_minimal() +
  theme(
    plot.background = element_blank(),
    plot.title = element_text(size = 12),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.justification = c(0, 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),
    strip.text = element_text(size = 10),
    plot.margin = unit(c(2, 1, 1, 1), "cm")
  ) +
  labs(
    x = "Age (CE)",
    y = "Proportion of Community Trophic Structures",
    title = "Proportions of Community Trophic Structures Over Time by Lake"
  )

# ------------------------------------------------------------------------------
# Render final plot (warning suppressed due to known density dropouts)
# ------------------------------------------------------------------------------
suppressWarnings(print(plot_density_fclusts_time_lines_1B))

```
## Regional proportions trophic structures 
```{r fig.asp = 0.5, fig.width = 12}
# Create the plot
plot_stacked_fclusts_time_lines <- 
  ls_all_hclusts_complete_paleo %>%
  map(~.x %>% as.data.frame()) %>% 
  bind_rows() %>%
  ggplot(aes(x = age_ce, fill = amd_clusts)) +
  geom_density(adjust = 1.5, position = "fill", colour = "white") +
  scale_fill_viridis_d(name = "Community Trophic \nstructures", option = "viridis", direction = -1) +
  lims(x = c(0, 2010)) +
  # Add vertical lines for all highlighted dates
  geom_vline(data = data.frame(x = highlighted_dates), aes(xintercept = x), colour = "white") +
  # Add labels for the dates
  geom_text(
    data = data.frame(x = highlighted_dates, y = 1.05, label = as.character(highlighted_dates)),
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    colour = "black",
    size = 3,
    fontface = "bold"
  ) +
  theme(plot.background = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.justification = c(0, 1),
    # legend.position = c(0, 1),
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.major = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Age (CE)",
    y = "Proportion of Community Trophic Structures",
    title = "Combined proportions of Community Trophic Structures over Time"
  )

plot_stacked_fclusts_time_lines    

bin_size <- 30

df_fclusts <- ls_all_hclusts_complete_paleo %>%
  purrr::map_dfr(~as.data.frame(.x)) %>%
  filter(!is.na(age_ce), !is.na(amd_clusts), !is.na(lake)) %>%
  mutate(
    amd_clusts = factor(as.character(amd_clusts)),   # stable factor
    age_bin    = floor(age_ce / bin_size) * bin_size
  ) %>%
  group_by(lake, age_bin) %>%
  mutate(w = 1 / n()) %>%         # each lake contributes total weight 1 per bin
  ungroup()

plot_regional_lake_balanced <-
  ggplot(df_fclusts, aes(x = age_ce, fill = amd_clusts)) +
  geom_density(aes(weight = w),
               adjust = 1.5, position = "fill", colour = "white") +
  scale_fill_viridis_d(name = "Community Trophic \nstructures",
                       option = "viridis", direction = -1) +
  lims(x = c(0, 2010)) +
  geom_vline(data = data.frame(x = highlighted_dates),
             aes(xintercept = x), colour = "white") +
  geom_text(
    data = data.frame(x = highlighted_dates, y = 1.05, label = as.character(highlighted_dates)),
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    colour = "black", size = 3, fontface = "bold"
  ) +
  theme_minimal() +
  labs(
    x = "Age (CE)",
    y = "Proportion of Community Trophic Structures",
    title = "Combined proportions of Community Trophic Structures over Time"
  )

plot_regional_lake_balanced

```


```{r fig.asp = 1, fig.width = 14}
# K values to compare
Ks <- c(2, 3, 4, 5, 6, 7, 8)

# coverage (how many lakes per age_bin)
cov <- df_fclusts %>%
  distinct(lake, age_bin) %>%
  count(age_bin, name = "n_lakes")

# build one df per K (and tag it)
df_common_allK <- map_dfr(Ks, function(K) {
  df_fclusts %>%
    left_join(cov, by = "age_bin") %>%
    filter(n_lakes >= K) %>%
    mutate(K = factor(K, levels = Ks))
})

# ensure stable legend ordering across facets
df_common_allK <- df_common_allK %>%
  mutate(amd_clusts = factor(amd_clusts, levels = sort(unique(df_fclusts$amd_clusts))))

# plot: 3 vertical panels
plot_common_support_Ks <-
  ggplot(df_common_allK, aes(x = age_ce, fill = amd_clusts)) +
  geom_density(
    aes(weight = w),
    adjust = 1.5,
    position = "fill",
    colour = "white"
  ) +
  facet_wrap(
    ~ K,
    ncol = 1,
    labeller = as_labeller(function(x)
      paste0("Common support: \u2265", x, " lakes"))
  ) +
  scale_fill_viridis_d(
    name = "Community Trophic \nstructures",
    option = "viridis",
    direction = -1
  ) +
  coord_cartesian(
    xlim = c(0, 2000),
    ylim = c(0, 1)
  ) +
  geom_vline(
    data = data.frame(x = highlighted_dates),
    aes(xintercept = x),
    colour = "white"
  ) +
  lims(x = c(0, 2000)) + 
  geom_text(
    data = data.frame(
      x = highlighted_dates,
      y = 1.05,
      label = as.character(highlighted_dates)
    ),
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    colour = "black",
    size = 3,
    fontface = "bold"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.background = element_rect(
      fill = "white",
      linewidth = 4,
      colour = "white"
    ),
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.major = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Age (CE)",
    y = "Proportion of Community Trophic Structures",
    title = "Regional trophic-structure proportions under common-support filtering (0–2000 CE)"
  )

plot_common_support_Ks

```

## Ordination with trophic structures
```{r fig.asp = 0.5, fig.width = 12}

# ------------------------------------------------------------------------------
# Compute ordination using PCoA on Bray-Curtis dissimilarity
# ------------------------------------------------------------------------------
ordinations_paleo <- norm_abund %>%
  as.data.frame() %>%
  select(-c(lake:age_ce)) %>%                        # Drop metadata columns
  vegdist(method = "bray", na.rm = TRUE) %>%         # Bray-Curtis distance
  cmdscale() %>%                                     # PCoA
  scores() %>%
  as.data.frame()

# ------------------------------------------------------------------------------
# Fit environmental vectors (alternative functional groups) to ordination
# ------------------------------------------------------------------------------
fit_env_paleo <- ordinations_paleo %>%
  envfit(
    env = select(norm_abund, alternative_functional_groups_guilds_globi_merged),
    scaling = "sites"
  ) %>%
  scores("vectors") %>%
  as.data.frame() %>%
  mutate(
    Dim1 = Dim1 * 0.75,
    Dim2 = Dim2 * 0.75,
    groups = rownames(.)
  )

# ------------------------------------------------------------------------------
# Compute convex hulls (polygons) for each trophic structure (cluster)
# ------------------------------------------------------------------------------
hull_paleo <- bind_cols(
    ordinations_paleo,
    select(norm_abund, alternative_functional_groups_guilds_globi_merged),
    select(global_matrix_merged_paleo, DCA1:amd_clusts)
  ) %>%
  group_by(amd_clusts) %>%
  dplyr::slice(chull(Dim1, Dim2))  # Convex hull per cluster

# ------------------------------------------------------------------------------
# Rename cluster levels for better readability (CTS1–CTS5)
# ------------------------------------------------------------------------------
global_matrix_merged_paleo$amd_clusts <- factor(
  global_matrix_merged_paleo$amd_clusts,
  levels = 1:5,
  labels = c("CTS1", "CTS2", "CTS3", "CTS4", "CTS5")
)

hull_paleo$amd_clusts <- factor(
  hull_paleo$amd_clusts,
  levels = 1:5,
  labels = c("CTS1", "CTS2", "CTS3", "CTS4", "CTS5")
)

# ------------------------------------------------------------------------------
# Assemble full ordination plot with polygons, points, and fitted vectors
# ------------------------------------------------------------------------------
ordinations_fclusts_1 <- ordinations_paleo %>%
  bind_cols(
    select(norm_abund, -alternative_functional_groups_guilds_globi_merged),
    .,
    select(global_matrix_merged_paleo, DCA1:amd_clusts)
  ) %>%
  ggplot(aes(x = Dim1, y = Dim2)) +
  
  # Polygons for each trophic structure
  geom_polygon(
    data = hull_paleo,
    aes(fill = as.factor(amd_clusts)),
    alpha = 0.2
  ) +
  scale_fill_viridis_d(option = "viridis", direction = -1, guide = "none") +
  ggnewscale::new_scale_fill() +
  
  # Points colored by cluster
  geom_point(aes(colour = as.factor(amd_clusts)), size = 3) +
  scale_colour_viridis_d(
    option = "viridis", direction = -1,
    name = "Community Trophic Structures"
  ) +
  ggnewscale::new_scale_colour() +

  # Environmental vectors
  geom_segment(
    data = fit_env_paleo,
    aes(x = 0, xend = Dim1, y = 0, yend = Dim2),
    arrow = arrow(length = unit(0.1, "cm"))
  ) +

  # Environmental labels
  ggrepel::geom_label_repel(
    data = fit_env_paleo,
    aes(x = Dim1, y = Dim2, label = groups),
    max.overlaps = 12,
    size = 3,
    direction = "both",
    segment.size = 0.25
  ) +

  # Styling
  theme_bw() +
  labs(title = "") +
  theme(
    plot.title = element_text(size = 12),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.justification = c(0, 1),
    legend.position = NULL,
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.major = element_line(colour = "grey70", linewidth = 0.2),
    panel.grid.minor = element_blank()
  ) +
  guides(fill = "none")

# ------------------------------------------------------------------------------
# Render plot
# ------------------------------------------------------------------------------
ordinations_fclusts_1

```

## Random Forest
```{r}

# ----------------------------
# Data Preparation
# ----------------------------

# Combine global matrix with normalized abundance data (excluding lake to age_ce columns)
datRF_paleo <- bind_cols(
  global_matrix_merged_paleo,
  select(norm_abund, -c(lake:age_ce))
) %>% 
  select(amd_clusts:last_col()) %>%     # Keep columns from amd_clusts to the last
  select(-total_nspp_by_lake_core)      # Remove total species richness column

# ----------------------------
# Random Forest Classification
# ----------------------------

# Convert cluster variable to factor for classification
datRF_paleo$amd_clusts <- as.factor(datRF_paleo$amd_clusts)

# Display class distribution
table(datRF_paleo$amd_clusts)

# Get the minimum class size (for potential balancing use)
min <- as.numeric(min(table(datRF_paleo$amd_clusts)))

# Fit random forest classifier with 1000 trees and variable importance enabled
RFfit_paleo <- randomForest(amd_clusts ~ ., data = datRF_paleo, ntree = 1000, importance = TRUE)

# ----------------------------
# 1.1 Out-of-Bag (OOB) Prediction Accuracy
# ----------------------------

# Predict using OOB samples (internal validation)
pred_paleo <- predict(RFfit_paleo)

# Append predictions to the original data
datRF_paleo <- cbind(pred_paleo, datRF_paleo)

# Confusion matrix: observed vs predicted clusters
tabrp_paleo <- table(datRF_paleo$amd_clusts, datRF_paleo$pred_paleo)

# Print row-wise classification accuracy (%)
round(100 * (prop.table(tabrp_paleo, 1)), 0)

# Print overall classification accuracy
paste("Classification accuracy:", round((100 * (sum(diag(tabrp_paleo)) / sum(tabrp_paleo))), 0), "%")

# Compute Cohen’s Kappa for classification agreement
kappa <- kappa2(datRF_paleo[, c(1, 2)], "equal")
round(kappa$value, 2)  # Interpret as moderate if between 0.4 and 0.6

# Diagonal: per-class correct classification %
diag(round(100 * (prop.table(tabrp_paleo, 1)), 0))

# ----------------------------
# Variable Importance and Partial Dependence Profiles (PDPs)
# ----------------------------

# Use DALEX to explain model behavior
explainer_paleo <- explain(model = RFfit_paleo, datRF_paleo[, -c(1:2)], datRF_paleo$amd_clusts)

# Compute and display variable importance via DALEX
imp_paleo <- variable_importance(explainer_paleo)
imp_paleo
plot(imp_paleo)

# Compute and plot PDPs (marginal effects of predictors)
pdp10_paleo <- model_profile(explainer_paleo)
plot(pdp10_paleo)

# ----------------------------
# Decision Tree Visualization
# ----------------------------

# Re-confirm amd_clusts is treated as a factor
datRF_paleo <- datRF_paleo %>%
  mutate(amd_clusts = as.factor(amd_clusts))

# Fit a decision tree using rpart (excluding predictions and response itself)
decision_tree_paleo <- rpart(
  datRF_paleo$amd_clusts ~ .,
  data = select(datRF_paleo, -c(pred_paleo, amd_clusts)),
  method = "class",
  cp = 0.035
)

# Display complexity parameter table
printcp(decision_tree_paleo)

# --------------------------------------------
# Save the decision tree as a high-resolution PNG image
# --------------------------------------------
png("outputs/decision_tree_paleo.png", width = 2000, height = 1400, res = 300)
rpart.plot::rpart.plot(
  decision_tree_paleo, 
  type = 5, 
  extra = 2, 
  legend.x = NA, 
  box.palette = as.list(viridis::viridis(n = ls_opt_clusts, direction = -1, alpha = 0.75))
)
dev.off()

```

# Analysis - Hierarchical Generalized Additive Models - HGAMs
## GAM - Diatoms
### Regional Smooth
```{r}

lakes_col <- c(viridis::viridis(9, option = "plasma", alpha = 0.8))
lakes_col_10 <- c("black", viridis::viridis(10, option = "plasma", alpha = 0.8))

b_ls_dca_diatoms <-
  b_ls_dca_diatoms %>%
  dplyr::filter(!is.na(age_ce)) %>% 
  dplyr::filter(!lake %in% c("Fogo", "Furnas" ))

# global model diatoms
modGI_diatoms <- 
  b_ls_dca_diatoms %>%
  mutate(lake = as.factor(lake)) %>%
  gam(
    DCA1 ~
      s(age_ce, bs = "tp", k = 30) +
      s(age_ce, by = lake, k = 8, m = 1, bs = "tp") +
      s(lake, bs = "re", k = 8),
    data = ., method = "REML"
  )

# Output p.Time.GI, p.time.derivative
mod_diatoms <- modGI_diatoms

# Model diagnostics
summary(mod_diatoms)
k.check(mod_diatoms, subsample=5000, n.rep=400)
gam.check(mod_diatoms)
appraise(mod_diatoms)
gratia::draw(mod_diatoms, residuals = T)

# get the model data to plot
# evaluate the smooths
smooth_diatoms <- smooth_estimates(mod_diatoms) %>%
  add_confint()

# add partial residuals to data
df_partial_residuals_diatoms <- 
  b_ls_dca_diatoms %>%
  add_partial_residuals(., model = mod_diatoms)

# add label
sum1 <- summary(mod_diatoms)
lab1 <- paste("AdjRsq = ", round(sum1$r.sq,2),
              "\nDevExpl = ", round(sum1$dev.expl,2))

plot_time_GI_diatoms <- 
  smooth_diatoms %>%
  filter(.smooth == "s(age_ce)") %>%
  ggplot() +
  geom_rug(aes(x = age_ce),
    data = df_partial_residuals_diatoms,
    sides = "b", length = grid::unit(0.02, "npc")
  ) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = age_ce),
    alpha = 0.2
  ) +
  # geom_point(aes(x = age_ce, y = `s(age_ce)`, color = lake),
  #           data = df_partial_residuals_diatoms, alpha = 0.5, size = 1) +
  geom_line(aes(x = age_ce, y = .estimate), lwd = 1) +
  labs(y = "DCA1", title = "s(age_ce) GI model") +
  scale_color_manual("Time-Interval") +
  theme_classic(base_size = 12) +
  scale_x_continuous(n.breaks = 5, limits = c(0, 2020)) +
  lims(y = c(-2, 2)) +
  labs(title = NULL) +
  theme(panel.grid.minor.y = element_blank()) +
  theme(legend.position = "none") +
    annotate("text_npc", npcx = 0.05, npcy = .95, label = lab1, cex = 3.5)

#plot_time_GI_diatoms

```

## GAM - Chironomids
### Regional Smooth

```{r}

b_ls_dca_chiro <-
  b_ls_dca_chiro %>%
  dplyr::filter(!is.na(age_ce)) %>% 
  dplyr::filter(!lake %in% c("Fogo", "Furnas" ))

# global model chiro
modGI_chiro <- 
  b_ls_dca_chiro %>%
  mutate(lake = as.factor(lake)) %>%
  gam(
    DCA1 ~
      s(age_ce, bs = "tp", k = 30) +
      s(age_ce, by = lake, k = 8, m = 1, bs = "tp") +
      s(lake, bs = "re", k = 8),
    data = ., method = "REML"
  )

# Output p.Time.GI, p.time.derivative
mod_chiro <- modGI_chiro

# Model diagnostics
summary(mod_chiro)
k.check(mod_chiro, subsample=5000, n.rep=400)
gam.check(mod_chiro)
appraise(mod_chiro)
gratia::draw(mod_chiro, residuals = T)

# get the model data to plot
# evaluate the smooths
smooth_chiro <- smooth_estimates(mod_chiro) %>%
  add_confint()

# add partial residuals to data
df_partial_residuals_chiro <- 
  b_ls_dca_chiro %>%
  add_partial_residuals(., model = mod_chiro)

# add label
sum1 <- summary(mod_chiro)
lab1 <- paste("AdjRsq = ", round(sum1$r.sq,2),
              "\nDevExpl = ", round(sum1$dev.expl,2))

plot_time_GI_chiro <- 
  smooth_chiro %>%
  filter(.smooth == "s(age_ce)") %>%
  ggplot() +
  geom_rug(aes(x = age_ce),
    data = df_partial_residuals_chiro,
    sides = "b", length = grid::unit(0.02, "npc")
  ) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = age_ce),
    alpha = 0.2
  ) +
  # geom_point(aes(x = age_ce, y = `s(age_ce)`, color = lake),
  #           data = df_partial_residuals_chiro, alpha = 0.5, size = 1) +
  geom_line(aes(x = age_ce, y = .estimate), lwd = 1) +
  labs(y = "DCA1", title = "s(age_ce) GI model") +
  scale_color_manual("Time-Interval", values = lakes_col) +
  theme_classic(base_size = 12) +
  scale_x_continuous(n.breaks = 5, limits = c(0, 2020)) +
  lims(y = c(-2, 2)) +
  labs(title = NULL) +
  theme(panel.grid.minor.y = element_blank()) +
  theme(legend.position = "none") +
    annotate("text_npc", npcx = 0.05, npcy = .95, label = lab1, cex = 3.5)

#plot_time_GI_chiro

```
## GAM - Fgroups
### Regional Smooth

```{r}

b_ls_dca_fgroup <-
  b_ls_dca_fgroup %>%
  dplyr::filter(!is.na(age_ce)) %>% 
  dplyr::filter(!lake %in% c("Fogo", "Furnas" ))

# global model fgroup
modGI_fgroup <- 
  b_ls_dca_fgroup %>%
  mutate(lake = as.factor(lake)) %>%
  gam(
    DCA1 ~
      s(age_ce, bs = "tp", k = 30) +
      s(age_ce, by = lake, k = 8, m = 1, bs = "tp") +
      s(lake, bs = "re", k = 8),
    data = ., method = "REML"
  )

# Output p.Time.GI, p.time.derivative
mod_fgroup <- modGI_fgroup

# Model diagnostics
summary(mod_fgroup)
k.check(mod_fgroup, subsample=5000, n.rep=400)
gam.check(mod_fgroup)
appraise(mod_fgroup)
gratia::draw(mod_fgroup, residuals = T)

# get the model data to plot
# evaluate the smooths
smooth_fgroup <- smooth_estimates(mod_fgroup) %>%
  add_confint()

# add partial residuals to data
df_partial_residuals_fgroup <- 
  b_ls_dca_fgroup %>%
  add_partial_residuals(., model = mod_fgroup)

# add label
sum1 <- summary(mod_fgroup)
lab1 <- paste("AdjRsq = ", round(sum1$r.sq,2),
              "\nDevExpl = ", round(sum1$dev.expl,2))

plot_time_GI_fgroup <- 
  smooth_fgroup %>%
  filter(.smooth == "s(age_ce)") %>%
  ggplot() +
  geom_rug(aes(x = age_ce),
    data = df_partial_residuals_fgroup,
    sides = "b", length = grid::unit(0.02, "npc")
  ) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = age_ce),
    alpha = 0.2
  ) +
  # geom_point(aes(x = age_ce, y = `s(age_ce)`, color = lake),
  #           data = df_partial_residuals_fgroup, alpha = 0.5, size = 1) +
  geom_line(aes(x = age_ce, y = .estimate), lwd = 1) +
  labs(y = "DCA1", title = "s(age_ce) GI model") +
  scale_color_manual("Time-Interval", values = lakes_col) +
  theme_classic(base_size = 12) +
  scale_x_continuous(n.breaks = 5, limits = c(0, 2020)) +
  lims(y = c(-2, 2)) +
  labs(title = NULL) +
  theme(panel.grid.minor.y = element_blank()) +
  theme(legend.position = "none") +
    annotate("text_npc", npcx = 0.05, npcy = .95, label = lab1, cex = 3.5)

#plot_time_GI_fgroup

```

# Multivariate Analysis - Redundancy Analysis and Variance Partitioning
## Data prep - Vegetation change - Regional Vegetation change
```{r}
# ----------------------------------------------------------
# Load and clean pollen data
# ----------------------------------------------------------

pollen <- pollen %>% 
  rename(age_bp = age) %>% 
  mutate(age_ce = 1950 - age_bp) %>% 
  mutate(ID = paste0(sitename, "_", age_ce))

df_1 <- pollen

# Filter out aquatic and unidentified taxa, then group by ecological group
df_wk <- df_1 %>%
  filter(!ecologicalgroup %in% c("UNID", "AQVP", "AQBR")) %>%
  group_by(ID, sitename, age_ce, group) %>%
  summarise(value = sum(value), .groups = 'drop') %>%
  group_by(ID) %>%
  mutate(percentage = value / sum(value) * 100) %>%
  ungroup()

# Calculate group-wise mean percentages and anomalies
df_wk <- df_wk %>%
  group_by(sitename, group) %>%
  mutate(mean_percentage = mean(percentage)) %>%
  ungroup() %>%
  mutate(percent_anomaly = percentage - mean_percentage) %>%
  select(-mean_percentage) %>%
  droplevels()

sites_col <- scico(n_distinct(df_wk$sitename), begin = 0, end = 0.95, palette = 'roma')
gr_col <- scico(n_distinct(df_wk$group), begin = 0, end = 0.95, palette = 'nuuk')

# ----------------------------------------------------------
# DCA Analysis
# ----------------------------------------------------------

exclude_sites <- c("Lomba", "Rasa")
df_wk_dca <- df_1 %>%
  filter(!ecologicalgroup %in% c("UNID"), age_ce > 0, !sitename %in% exclude_sites) %>%
  droplevels()

results_list <- lapply(sort(unique(df_wk_dca$sitename)), function(site) {
  df.sp <- df_wk_dca %>%
    filter(sitename == site) %>%
    group_by(ID, taxa) %>%
    summarise(value = sum(value), .groups = 'drop') %>%
    pivot_wider(names_from = taxa, values_from = value, values_fill = 0) %>%
    column_to_rownames("ID") %>%
    select(where(~ any(. != 0)))
  
  df.spe <- decostand(df.sp, method = "hellinger")
  df.DCA <- decorana(df.spe) %>%
    scores(choices = 1) %>%
    as.data.frame() %>%
    rownames_to_column("ID")
  
  return(df.DCA)
})

DCA.res <- Reduce(full_join, results_list) %>%
  separate(ID, into = c("sitename", "age_ce"), sep = "_", convert = TRUE) %>%
  arrange(sitename, age_ce) %>%
  mutate(DCA1 = case_when(
    sitename == "Azul" ~ -DCA1,
    sitename == "Gingal" ~ -DCA1,
    TRUE ~ DCA1
  ))

DF_to <- DCA.res %>% mutate(sitename = as.factor(sitename))

# ----------------------------------------------------------
# HGAM on DCA
# ----------------------------------------------------------

mod_to <- gam(DCA1 ~ s(age_ce, bs = "tp", k = 25) +
                s(age_ce, by = sitename, bs = "tp", k = 15, m = 1) +
                s(sitename, bs = "re"),
              data = DF_to, method = "REML")

summary(mod_to)
k.check(mod_to, n.rep = 500)
appraise(mod_to)

# Plot smoothed global trend and derivative
sm_to <- smooth_estimates(mod_to) %>% add_confint()
df <- DF_to %>% add_partial_residuals(mod_to)
lakes_col <- scico(length(levels(DF_to$sitename)), begin = 0, end = 0.95, palette = 'roma')

p.DCA.gl <- sm_to %>%
  filter(.smooth == "s(age_ce)") %>%
  ggplot() +
  geom_rug(aes(x = age_ce), data = df, sides = "b", length = unit(0.02, "npc")) +
  geom_ribbon(aes(x = age_ce, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_point(aes(x = age_ce, y = `s(age_ce)`, color = sitename), data = df, alpha = 0.8, size = 1.5) +
  geom_line(aes(x = age_ce, y = .estimate), linewidth = 1) +
  labs(y = "DCA axis-1", x = "Age (CE)") +
  scale_color_manual("Lake", values = lakes_col) +
  theme_minimal(base_size = 9) +
  scale_x_continuous(n.breaks = 10) +
  theme(panel.grid.minor.y = element_blank(), legend.position = "bottom")

mod_to_dev <- derivatives(mod_to, select = "s(age_ce)", eps = 1, level = 0.9, interval = "simultaneous", seed = 22)
color_mapping <- scico(2, begin = 0, end = 0.9, palette = 'vik')

# ----------------------------------------------------------
# HGAM on representative groups
# ----------------------------------------------------------

pollen_clean <- pollen %>%
  filter(elementtype == "pollen", !is.na(value)) %>%
  filter(!sitename %in% c("Lomba", "Rasa"))

sample_totals <- pollen_clean %>%
  group_by(sitename, age_ce) %>%
  summarise(sample_total = sum(value, na.rm = TRUE), .groups = "drop")

lake_avg <- sample_totals %>%
  group_by(sitename) %>%
  summarise(avg_lake_total = mean(sample_total, na.rm = TRUE), .groups = "drop")

pollen_normalized <- pollen_clean %>%
  left_join(sample_totals, by = c("sitename", "age_ce")) %>%
  left_join(lake_avg, by = "sitename") %>%
  mutate(
    scaling_factor = avg_lake_total / sample_total,
    value_scaled = value * scaling_factor
  )

# --- Calculate percent composition by group ---

pollen_percent_group <- pollen_normalized %>%
  group_by(sitename, age_ce, group) %>%
  summarise(value_scaled = sum(value_scaled, na.rm = TRUE), .groups = "drop") %>%
  group_by(sitename, age_ce) %>%
  mutate(
    total_scaled = sum(value_scaled),
    percent = (value_scaled / total_scaled) * 100
  ) %>%
  filter(!is.na(percent)) %>%
  mutate(group = factor(group), sitename = factor(sitename)) %>%
  ungroup()

pollen_percent_group <- pollen_percent_group %>%
  mutate(sitename = factor(sitename))  # ensure factor type

# --- GAM fitting function ---
#   defined in `functions/source_custom_functions.R` (fit_gam_group)

# --- Fit GAM for each pollen group ---

gam_models <- pollen_percent_group %>%
  group_split(group) %>%
  set_names(map_chr(., ~ as.character(unique(.x$group)))) %>%
  map(fit_gam_group)

# --- Plotting function per group ---
#   defined in `functions/source_custom_functions.R` (plot_gam_group)

# --- Generate plots ---

plots_all <- map2(names(gam_models), gam_models, plot_gam_group)

final_plot <- wrap_plots(plots_all, ncol = 2) +
  plot_annotation(
    title = "% of Pollen Groups Over Time (GAM fit)",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

# --- Display plot ---

final_plot

# --- Compare tree vs herb pollen groups ---

tree_data <- pollen_percent_group %>% filter(group == "Tree")
herb_data <- pollen_percent_group %>% filter(group == "Herb")

mod_tree <- fit_gam_group(tree_data)
mod_herb <- fit_gam_group(herb_data)

aug_tree <- augment(mod_tree, type.predict = "response") %>%
  mutate(group = "Tree")

aug_herb <- augment(mod_herb, type.predict = "response") %>%
  mutate(group = "Herb")

combined_df_pollen <- bind_rows(aug_tree, aug_herb)
```

### Pollen analysis
```{r fig.width = 12, fig.asp = 0.5}
# --- taxa lists (edit freely) ---
taxa_cereals   <- c("Cerealia", "Triticum", "Secale", "Zea mays")
taxa_plantago  <- c("Plantago", "Plantago lanceolata", "Plantago major", "Plantago coronopus")

# Conservative "ruderal" list (expand if you want)
taxa_ruderal   <- c("Amaranthaceae", "Artemisia", "Carduus",
                    "Xanthium", "Xanthium strumarium", "Ambrosia",
                    "Achillea", "Centaurea cyanus", "Urtica", "Urtica urens")

# Native/endemic Azores indicators from your taxa list (expand as desired)
taxa_native <- c(
  "Juniperus brevifolia", "Picconia azorica", "Frangula azorica",
  "Prunus lusitanica subsp. azorica", "Prunus lusitanica",
  "Myrica faya", "Erica azorica", "Daboecia azorica",
  "Viburnum treleasei", "Ilex perado subsp. azorica",
  "Hypericum foliosum", "Hedera azorica", "Chaerophyllum azoricum",
  "Sanicula azorica", "Daucus carota subsp. azoricus",
  "Lysimachia azorica", "Vicia dennesiana", "Arceuthobium azoricum"
)

bin_size <- 30

# highlighted dates (your vector)
# highlighted_dates <- c(750, 1050, 1450, 1750)

# ------------------------------------------------------------
# Denominator: terrestrial pollen total (scaled), per sample
# ------------------------------------------------------------
sample_totals_scaled <- pollen_normalized %>%
  filter(
    elementtype == "pollen",
    !is.na(value_scaled),
    age_ce > 0,
    !ecologicalgroup %in% c("UNID", "AQVP", "AQBR")
  ) %>%
  group_by(sitename, age_ce) %>%
  summarise(total_scaled = sum(value_scaled, na.rm = TRUE), .groups = "drop")

# ------------------------------------------------------------
# Numerators: sum of scaled pollen for each category, per sample
# ------------------------------------------------------------
pollen_4cat <- pollen_normalized %>%
  filter(
    elementtype == "pollen",
    !is.na(value_scaled),
    age_ce > 0,
    !ecologicalgroup %in% c("UNID", "AQVP", "AQBR")
  ) %>%
  mutate(
    category = case_when(
      taxa %in% taxa_cereals  ~ "Cereals (Cerealia)",
      taxa %in% taxa_plantago ~ "Plantago",
      taxa %in% taxa_ruderal  ~ "Ruderal weeds",
      taxa %in% taxa_native   ~ "Native pollen",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(category)) %>%
  group_by(sitename, age_ce, category) %>%
  summarise(value_scaled_cat = sum(value_scaled, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_totals_scaled, by = c("sitename", "age_ce")) %>%
  mutate(
    percent = 100 * value_scaled_cat / total_scaled,
    age_bin = floor(age_ce / bin_size) * bin_size,
    category = factor(category, levels = c("Native pollen", "Cereals (Cerealia)", "Plantago", "Ruderal weeds"))
  ) %>%
  filter(is.finite(percent))

# Optional: thin bins for plotting (every 60 yrs)
pollen_4cat_plot <- pollen_4cat %>%
  mutate(age_bin60 = floor(age_ce / 60) * 60)

# ------------------------------------------------------------
# Colors: native = green, others = yellow→orange→red
# ------------------------------------------------------------
cat_fill <- c(
  "Native pollen"       = "#2E8B57",  # green
  "Cereals (Cerealia)"  = "#F4D35E",  # yellow
  "Plantago"            = "#F08A24",  # orange
  "Ruderal weeds"       = "#D1495B"   # red
)

# ------------------------------------------------------------
# Violin plot through time + highlighted dates (vertical lines)
# ------------------------------------------------------------
p_violin_time <- ggplot(
  pollen_4cat_plot,
  aes(x = age_bin60, y = percent)
) +

  # --- Historical phase markers ---
  geom_vline(
    xintercept = highlighted_dates,
    linewidth = 0.5,
    linetype = "dashed",
    alpha = 0.6
  ) +

  # --- BACKGROUND POINTS (show singletons & rare events) ---
  geom_point(
    aes(color = category),
    position = position_jitter(width = 10, height = 0),
    alpha = 0.25,
    size = 0.9
  ) +

  # --- VIOLINS (only where density exists) ---
  geom_violin(
    aes(
      fill  = category,
      group = interaction(category, age_bin60)
    ),
    scale = "width",
    adjust = 1,
    trim = TRUE,
    alpha = 0.75
  ) +

  facet_wrap(~ category, ncol = 1, scales = "free_y") +

  scale_fill_manual(values = cat_fill) +
  scale_color_manual(values = cat_fill) +

  labs(
    x = "Age (CE) (60-year bins for plotting)",
    y = "% of terrestrial pollen (scaled within lake)",
    title = "Anthropogenic and native pollen indicators through time",
    subtitle = "Points show all observations; violins indicate distributions where sufficient data exist"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

p_violin_time

# Optional: sqrt scale version (helps when magnitudes differ a lot)
p_violin_time_sqrt <- p_violin_time +
  scale_y_continuous(trans = "sqrt") +
  labs(
    y = "% of terrestrial pollen (scaled within lake) [sqrt scale]",
    subtitle = "Violin distributions per time bin (across lakes/samples); dashed lines = highlighted dates; sqrt y-scale"
  )

p_violin_time_sqrt
```
### Pollen analysis
```{r fig.width = 14, fig.asp = 0.5}
# ------------------------------------------------------------
# ASSUMES you already have:
#  - pollen_4cat_plot (includes age_bin60, percent, category)
#  - sm_to (global vegetation smooth with age_ce + .estimate + CI columns)
#  - highlighted_dates (e.g., c(750,1050,1450,1750))
# ------------------------------------------------------------

# 1) Order groups bottom -> top (as requested)
cat_levels <- c("Native pollen", "Ruderal weeds", "Plantago", "Cereals (Cerealia)")

pollen_band <- pollen_4cat_plot %>%
  mutate(
    category = factor(category, levels = cat_levels),
    band_id  = as.integer(category) - 1L   # 0..3
  ) %>%
  group_by(category) %>%
  mutate(
    # Scale percent within each category so each band has comparable height
    percent_scaled = scales::rescale(percent, to = c(0.05, 0.95), na.rm = TRUE),
    y_band = band_id + percent_scaled
  ) %>%
  ungroup()

# 2) Single global GAM curve (scaled once to the full vertical span of bands)
veg_bg <- sm_to %>%
  dplyr::filter(.smooth == "s(age_ce)") %>%
  dplyr::transmute(
    age_ce  = age_ce,
    veg_est = .estimate,
    veg_lo  = .lower_ci,
    veg_hi  = .upper_ci
  ) %>%
  arrange(age_ce)

# Store original GAM scale for the RIGHT axis
veg_min <- min(veg_bg$veg_est, na.rm = TRUE)
veg_max <- max(veg_bg$veg_est, na.rm = TRUE)

# Scale GAM y to match band space (single mapping for all bands)
ymin_band <- min(pollen_band$y_band, na.rm = TRUE)
ymax_band <- max(pollen_band$y_band, na.rm = TRUE)

veg_bg <- veg_bg %>%
  mutate(
    veg_y     = scales::rescale(veg_est, to = c(ymin_band, ymax_band)),
    veg_lo_y  = scales::rescale(veg_lo,  to = c(ymin_band, ymax_band)),
    veg_hi_y  = scales::rescale(veg_hi,  to = c(ymin_band, ymax_band)),
    veg_est_scaled = scales::rescale(veg_est, to = c(0, 1))
  )

# 3) Colors: native green, others yellow→orange→red
cat_fill <- c(
  "Native pollen"       = "#2E8B57",
  "Cereals (Cerealia)"  = "#F4D35E",
  "Plantago"            = "#F08A24",
  "Ruderal weeds"       = "#D1495B"
)

# Helper: inverse mapping from band y-space -> original GAM space (for right axis)
#   defined in `functions/source_custom_functions.R` (inv_gam_from_band)

# 4) Build the plot (single GAM behind everything + RIGHT y-axis for GAM)
p_collage_singleGAM <- ggplot() +

  # --- single GAM ribbon (behind everything) ---
  geom_ribbon(
    data = veg_bg,
    aes(x = age_ce, ymin = veg_lo_y, ymax = veg_hi_y),
    inherit.aes = FALSE,
    fill = "grey70", alpha = 0.20
  ) +

  # --- single GAM line: thick + green→yellow gradient, transparent ---
  geom_path(
    data = veg_bg,
    aes(x = age_ce, y = veg_y, colour = veg_est_scaled),
    inherit.aes = FALSE,
    linewidth = 10,
    alpha = 0.30,
    lineend  = "butt",   # removes rounded caps
    linejoin = "mitre"   # reduces “bubble” joins
  ) +
  scale_colour_gradient(low = "#2E8B57", high = "#F4D35E", guide = "none") +

  ggnewscale::new_scale_colour() +

  # --- phase markers ---
  geom_vline(
    xintercept = highlighted_dates,
    linewidth = 0.5, linetype = "dashed", alpha = 0.6
  ) +

  # --- points (always shown, includes singletons) ---
  geom_point(
    data = pollen_band,
    aes(x = age_bin60, y = y_band, colour = category),
    position = position_jitter(width = 10, height = 0),
    alpha = 0.25, size = 0.9
  ) +

  # --- violins (only where density exists; singletons still visible as points) ---
  geom_violin(
    data = pollen_band,
    aes(
      x = age_bin60, y = y_band,
      fill  = category,
      group = interaction(category, age_bin60)
    ),
    scale = "width", adjust = 1, trim = TRUE, alpha = 0.75
  ) +

  scale_fill_manual(values = cat_fill) +
  scale_color_manual(values = cat_fill) +

  # --- LEFT axis: bands; RIGHT axis: GAM (secondary axis) ---
  scale_y_continuous(
    breaks = 0:3 + 0.5,
    labels = cat_levels,
    expand = expansion(mult = c(0.03, 0.03)),
    sec.axis = sec_axis(
      trans = inv_gam_from_band,
      name  = "Regional vegetation change (HGAM global smooth)"
    )
  ) +

  labs(
    x = "Age (CE) (60-year bins for plotting)",
    y = NULL,
    title = "Regional vegetation change with anthropogenic/native pollen indicators",
    subtitle = "Background: HGAM global vegetation smooth (± CI). Overlay: pollen distributions per 60-year bin (scaled within each indicator)."
  ) +

  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(face = "bold"),
    axis.title.y.right = element_text(face = "bold"),
    axis.text.y.right = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

p_collage_singleGAM
```


#### Pollen analysis - Regional
```{r fig.width = 12, fig.asp = 0.35}
bin_plot <- 60

cat_levels <- c("Cereals (Cerealia)", "Plantago", "Native pollen")

cat_cols <- c(
  "Native pollen"       = "#2E8B57",
  "Plantago"            = "#F08A24",
  "Cereals (Cerealia)"  = "#F4D35E"
)

# Presence = at least one lake with non-zero pollen in that bin
presence_by_bin <- pollen_4cat %>%
  filter(category %in% cat_levels) %>% 
  mutate(
    age_bin60 = floor(age_ce / bin_plot) * bin_plot,
    category  = factor(category, levels = cat_levels)
  ) %>%
  group_by(category, age_bin60, sitename) %>%
  summarise(present = any(percent > 0, na.rm = TRUE), .groups = "drop") %>%
  group_by(category, age_bin60) %>%
  summarise(
    n_lakes_present = sum(present),
    .groups = "drop"
  ) %>%
  filter(n_lakes_present > 0) %>%
  mutate(presence = 1)

p_presence_rows <- ggplot(
  presence_by_bin,
  aes(x = age_bin60, y = presence)
) +

  # Historical phase markers
  geom_vline(
    xintercept = highlighted_dates,
    linetype = "dashed",
    linewidth = 0.4,
    alpha = 0.6
  ) +

  # Presence markers
  geom_point(
    aes(size = n_lakes_present, color = category),
    alpha = 0.8
  ) +

  facet_wrap(
    ~ category,
    ncol = 1,
    scales = "free_y"
  ) +

  scale_color_manual(values = cat_cols) +
  scale_size_continuous(range = c(1.5, 4)) +

  scale_y_continuous(
    breaks = NULL,
    labels = NULL
  ) +

  labs(
    x = "Age (CE) (60-year bins)",
    y = NULL,
    title = "Regional appearance of anthropogenic and native pollen indicators",
    subtitle = "Each row shows when a pollen group is present in at least one lake; point size reflects number of lakes"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

p_presence_rows
```

#### Pollen analysis - by island
```{r fig.width = 12, fig.asp = 0.5}
lake_island_lut <- tribble(
  ~sitename,             ~island,
  "Azul",                "São Miguel",
  "Pico da Esperança",   "São Jorge",
  "Caldeirão",           "Corvo",
  "Caldeirão Connor",    "Corvo",
  "Caveiro",             "Pico",
  "Funda",               "Flores",
  "Gingal",              "Terceira",
  "Peixinho",            "Pico",
  "Pico Bog",            "Pico",
)

pollen_4cat2 <- pollen_4cat %>%
  left_join(lake_island_lut, by = "sitename")

# sanity checks
stopifnot("island" %in% names(pollen_4cat2))
if (anyNA(pollen_4cat2$island)) {
  print(unique(pollen_4cat2$sitename[is.na(pollen_4cat2$island)]))
  stop("Some sitenames are missing an island mapping (printed above).")
}

bin_plot <- 60

cat_levels <- c("Cereals (Cerealia)", "Plantago", "Native pollen")

cat_cols <- c(
  "Native pollen"       = "#2E8B57",
  "Plantago"            = "#F08A24",
  "Cereals (Cerealia)"  = "#F4D35E"
)

# Use the islands actually present (and keep a nice order)
island_levels <- c("Corvo", "Flores", "Terceira", "Pico", "São Miguel", "São Jorge")

presence_by_bin_island <- pollen_4cat2 %>%
  filter(category %in% cat_levels) %>% 
  mutate(
    age_bin60 = floor(age_ce / bin_plot) * bin_plot,
    category  = factor(category, levels = cat_levels),
    island    = factor(island, levels = island_levels)
  ) %>%
  group_by(category, island, age_bin60, sitename) %>%
  summarise(present = any(percent > 0, na.rm = TRUE), .groups = "drop") %>%
  group_by(category, island, age_bin60) %>%
  summarise(n_lakes_present = sum(present), .groups = "drop") %>%
  filter(n_lakes_present > 0)

presence_spans <- presence_by_bin_island %>%
  group_by(category, island) %>%
  summarise(
    age_start = min(age_bin60),
    age_end   = max(age_bin60),
    .groups = "drop"
  )

p_presence_rows_islands <- ggplot() +

  # Historical phase markers
  geom_vline(
    xintercept = highlighted_dates,
    linetype   = "dashed",
    linewidth  = 0.4,
    alpha      = 0.6
  ) +

  # Presence span (thick horizontal line)
  geom_segment(
    data = presence_spans,
    aes(
      x    = age_start,
      xend = age_end,
      y    = island,
      yend = island,
      color = category
    ),
    linewidth = 2.2,
    alpha = 0.6,
    lineend = "round"
  ) +

  # Discrete presence points
  geom_point(
    data = presence_by_bin_island,
    aes(
      x = age_bin60,
      y = island,
      size = n_lakes_present,
      color = category
    ),
    alpha = 0.9
  ) +

  facet_wrap(~ category, ncol = 1) +

  scale_color_manual(values = cat_cols) +
  scale_size_continuous(range = c(1.5, 5)) +

  labs(
    x = "Age (CE) (60-year bins)",
    y = NULL,
    title = "Regional appearance of anthropogenic and native pollen indicators",
    subtitle = "Horizontal bars indicate the temporal span of presence within each island; points show 60-year bins (size = number of lakes)"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

p_presence_rows_islands
```

#### Pollen analysis - by island, by lake
```{r fig.width = 12, fig.asp = 0.5}
bin_plot <- 60

cat_levels <- c("Cereals (Cerealia)", "Plantago", "Native pollen")

cat_cols <- c(
  "Native pollen"       = "#2E8B57",
  "Plantago"            = "#F08A24",
  "Cereals (Cerealia)"  = "#F4D35E"
)

# ---- one row per LAKE ----
presence_by_bin_lake <- pollen_4cat2 %>%
  filter(category %in% cat_levels) %>% 
  mutate(
    age_bin60 = floor(age_ce / bin_plot) * bin_plot,
    category  = factor(category, levels = cat_levels)
  ) %>%
  group_by(category, island, sitename, age_bin60) %>%
  summarise(present = any(percent > 0, na.rm = TRUE), .groups = "drop") %>%
  filter(present)  %>% 
  filter(!category %in% "Ruderal weeds")

# span per lake (first to last presence)
presence_spans_lake <- presence_by_bin_lake %>%
  group_by(category, island, sitename) %>%
  summarise(
    age_start = min(age_bin60),
    age_end   = max(age_bin60),
    .groups = "drop"
  )

# y-order: group lakes by island (nice visual blocks), then by sitename
lake_order <- presence_by_bin_lake %>%
  distinct(island, sitename) %>%
  arrange(island, sitename) %>%
  mutate(lake_label = paste0(island, " — ", sitename))

presence_by_bin_lake <- presence_by_bin_lake %>%
  left_join(lake_order, by = c("island", "sitename")) %>%
  mutate(lake_label = factor(lake_label, levels = lake_order$lake_label))

presence_spans_lake <- presence_spans_lake %>%
  left_join(lake_order, by = c("island", "sitename")) %>%
  mutate(lake_label = factor(lake_label, levels = lake_order$lake_label))

p_presence_rows_lakes <- ggplot() +

  # Historical phase markers
  geom_vline(
    xintercept = highlighted_dates,
    linetype   = "dashed",
    linewidth  = 0.4,
    alpha      = 0.6
  ) +

  # Presence span per lake (thick horizontal line)
  geom_segment(
    data = presence_spans_lake,
    aes(
      x = age_start, xend = age_end,
      y = lake_label, yend = lake_label,
      color = category
    ),
    linewidth = 2.2,
    alpha = 0.55,
    lineend = "round"
  ) +

  # Discrete presence points (one per bin per lake)
  geom_point(
    data = presence_by_bin_lake,
    aes(
      x = age_bin60,
      y = lake_label,
      color = category
    ),
    size = 2.2,
    alpha = 0.9
  ) +

  facet_wrap(~ category, ncol = 1) +
  scale_color_manual(values = cat_cols) +

  labs(
    x = "Age (CE) (60-year bins)",
    y = NULL,
    title = "Appearance of anthropogenic and native pollen indicators by lake",
    subtitle = "Horizontal bars show each lake’s presence span; points show 60-year bins with presence"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

p_presence_rows_lakes

```


## Data prep - Align environmental data
```{r fig.width = 12, fig.asp = 1.2}
# ------------------------------------------------------------------------------
# Vegetation smooth on a yearly grid WITHOUT interpolation
# ------------------------------------------------------------------------------

interp_years <- 0:2010
ref_site <- levels(DF_to$sitename)[1]

newdata_veg <- tibble(
  age_ce   = interp_years,
  sitename = factor(ref_site, levels = levels(DF_to$sitename))
)

pr_terms <- predict(mod_to, newdata = newdata_veg, type = "terms", se.fit = TRUE)

term_names <- colnames(pr_terms$fit)
idx <- which(term_names == "s(age_ce)")
if (length(idx) == 0) idx <- grep("^s\\(age_ce\\)$", term_names)
if (length(idx) != 1) {
  stop("Could not uniquely identify the global smooth term 's(age_ce)'.\n",
       "Available term names:\n", paste(term_names, collapse = ", "))
}

est <- as.numeric(pr_terms$fit[, idx])
se  <- as.numeric(pr_terms$se.fit[, idx])

alpha <- 0.05
zcrit <- qnorm(1 - alpha/2)

interpolated_df <- tibble(
  age_ce   = as.integer(interp_years),
  estimate = est,
  lower_ci = est - zcrit * se,
  upper_ci = est + zcrit * se
)

# ------------------------------------------------------------------------------
# Prepare NAO Hernandez (NO interpolation, NO binning)
# ------------------------------------------------------------------------------
colnames(nao_hernandez)[2] <- "age_ce"
colnames(nao_hernandez)[3] <- "nao_lower_ci"
colnames(nao_hernandez)[5] <- "NAO_Median_Value"
colnames(nao_hernandez)[7] <- "nao_upper_ci"

nao_hernandez_interpolated <- nao_hernandez %>%
  mutate(
    #age_ce = round(as.numeric(age_ce)),
    NAO_Median_Value = as.numeric(NAO_Median_Value)
  ) %>%
  filter(age_ce >= 0) %>%
  arrange(age_ce) %>%
  select(age_ce, NAO_Median_Value, nao_lower_ci, nao_upper_ci)

# ------------------------------------------------------------------------------
# Prepare NHST Büntgen (NO interpolation, NO binning)
# ------------------------------------------------------------------------------
nhst_buntgen <- nhst_buntgen %>%
  mutate(
    age_ce = round(as.numeric(age_ce)),
    Rmean  = as.numeric(Rmean)
  ) %>%
  filter(age_ce >= 0) %>%
  arrange(age_ce)

# ------------------------------------------------------------------------------
# Prepare normalized guild abundances
# ------------------------------------------------------------------------------
norm_abund_guilds <- norm_abund_guilds %>%
  mutate(age_ce = round(age_ce))

# ------------------------------------------------------------------------------
# CTS cluster proportions (requires purrr::map already loaded above)
# ------------------------------------------------------------------------------
combined_data <- ls_all_hclusts_complete_paleo %>%
  map(~as.data.frame(.x)) %>%
  bind_rows()

densities <- combined_data %>%
  group_by(amd_clusts) %>%
  do({
    dens <- density(.$age_ce, adjust = 1.5, from = 0, to = 2010, n = 5001)
    data.frame(age_ce = dens$x, density = dens$y)
  }) %>%
  ungroup()

proportions_long <- densities %>%
  group_by(age_ce) %>%
  mutate(total_density = sum(density),
         proportion = density / total_density) %>%
  ungroup() %>%
  mutate(age_ce = round(age_ce)) %>%
  filter(age_ce >= 0, age_ce <= 2010) %>%
  group_by(age_ce, amd_clusts) %>%
  summarise(average_proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

prop_comm_matrix <- proportions_long %>%
  tidyr::pivot_wider(names_from = amd_clusts, values_from = average_proportion)

# ------------------------------------------------------------------------------
# 30-year bin helper
# ------------------------------------------------------------------------------
bin_size <- 30

#   bin30() is defined in `functions/source_custom_functions.R`

# ------------------------------------------------------------
# 1) BIOLOGY: bin within lake first, then regional
# ------------------------------------------------------------
bio_lake_30 <- norm_abund_guilds %>%
  select(lake, age_ce, high_profile:predator) %>%
  filter(age_ce > 0) %>%
  mutate(age_ce = as.numeric(age_ce),
         age_bin = floor(age_ce / bin_size) * bin_size) %>%
  group_by(lake, age_bin) %>%
  summarise(across(high_profile:predator, ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  rename(age_ce = age_bin)

bio_reg_30 <- bio_lake_30 %>%
  group_by(age_ce) %>%
  summarise(
    n_lakes = n_distinct(lake),
    across(high_profile:predator, ~mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(age_ce)

# ------------------------------------------------------------
# 2) VEGETATION: bin once
# ------------------------------------------------------------
veg_30 <- bin30(interpolated_df, time_col = "age_ce",
                cols = c("estimate", "lower_ci", "upper_ci"))

# ------------------------------------------------------------
# 3) CLIMATE candidates: bin once
# ------------------------------------------------------------
nao_30          <- bin30(nao_hernandez_interpolated, time_col = "age_ce", cols = c("NAO_Median_Value"))
nhst_30         <- bin30(nhst_buntgen,               time_col = "age_ce", cols = c("Rmean"))

# ------------------------------------------------------------
# 4) CTS proportions: bin once
# ------------------------------------------------------------
prop_cols <- setdiff(names(prop_comm_matrix), "age_ce")
prop_comm_30 <- bin30(prop_comm_matrix, time_col = "age_ce", cols = prop_cols)

# ------------------------------------------------------------
# 5) FINAL JOIN: all binned tables share SAME age_ce grid
# ------------------------------------------------------------
joined_df_30yr <- bio_reg_30 %>%
  left_join(nao_30, by = "age_ce") %>%
  left_join(nhst_30, by = "age_ce") %>%                 # Büntgen NHST (Rmean)
  left_join(veg_30 %>% select(age_ce, estimate), by = "age_ce") %>%  # keep only estimate here
  left_join(prop_comm_30, by = "age_ce")

# ------------------------------------------------------------
# 6) Separate candidate table ONLY for preselection (no downstream dependency)
# ------------------------------------------------------------
env_candidates_30 <- bio_reg_30 %>%
  select(age_ce) %>%
  left_join(veg_30 %>% select(age_ce, estimate), by = "age_ce") %>%
  left_join(nao_30 %>% select(age_ce, NAO_Median_Value), by = "age_ce") %>%
  left_join(nhst_30 %>% select(age_ce, Rmean), by = "age_ce")

```

## Global RDA model
```{r fig.width = 8, fig.asp = 1}

# ============================================================
# GLOBAL RDA MODEL (consistent with varpart using NAO + vegetation)
# - Hellinger-transform response (as per Methods)
# - Screen predictors with ordiR2step (adjR2-based)
# - Keep a FINAL retained model with NAO + estimate for plotting + varpart
# ============================================================

# -----------------------------
# 1) Matrices
# -----------------------------
vars_to_test <- c(
  "NAO_Median_Value",
  "Rmean"#
)

comm_raw <- joined_df_30yr %>%
  dplyr::select(high_profile:predator)

env_raw <- joined_df_30yr %>%
  dplyr::select(age_ce, estimate, all_of(vars_to_test))

# keep complete cases across comm + env (same rows)
all_data <- bind_cols(comm_raw, env_raw)
complete_data <- all_data %>% filter(complete.cases(.))

comm <- complete_data %>% dplyr::select(high_profile:predator)
env  <- complete_data %>% dplyr::select(age_ce, estimate, all_of(vars_to_test))

# Hellinger transform (this is what your Methods claim)
comm_hel <- decostand(comm, method = "hellinger")

# All variables
explanatory <- env %>%
  dplyr::select(-age_ce)

# Marginal (single-predictor) tests: "significant on their own"
cat("\nMarginal (single-predictor) tests:\n")
marginal_res <- lapply(colnames(explanatory), function(v) {
  m <- capscale(comm_hel ~ ., data = explanatory[, v, drop = FALSE])
  a <- anova(m, permutations = 999)
  tibble(variable = v, F = a$F[1], p = a$`Pr(>F)`[1])
}) %>% bind_rows()

print(marginal_res)

explanatory <- env %>%
  dplyr::select(-age_ce, -Rmean, NAO_Median_Value, estimate)

final_model <- capscale(comm_hel ~ ., data = explanatory)

cat("\nFinal retained model (for biplot + matches varpart):\n")
print(anova(final_model, permutations = 999))

# -----------------------------
# 7) Biplot (only vectors for variables significant on their own OR retained)
# -----------------------------
# Criterion you asked earlier: show vectors for variables significant alone.
sig_alone <- marginal_res %>%
  filter(p < 0.05) %>%
  pull(variable)

# Always include final_vars (so climate is shown even if "alone" criterion shifts)
vars_to_plot <- unique(c(sig_alone))

bp_scores <- scores(final_model, display = "bp", scaling = 2)
bp_df <- as.data.frame(bp_scores) %>%
  rownames_to_column("Variable") %>%
  filter(Variable %in% vars_to_plot) %>%
  mutate(Label = recode(Variable,
                        estimate = "Veg. Change",
                        NAO_Median_Value = "NAO")
  )

# site scores (points)
site_scores <- scores(final_model, display = "sites", scaling = 2)
site_df <- as.data.frame(site_scores)
colnames(site_df)[1:2] <- c("CAP1", "CAP2")
site_df$age_ce <- env$age_ce

# species scores (guilds)
species_df <- as.data.frame(scores(final_model, display = "species", scaling = 2))
colnames(species_df)[1:2] <- c("CAP1", "CAP2")
species_df <- species_df %>%
  rownames_to_column("Guild") %>%
  mutate(Label = recode(Guild,
                        high_profile = "High Profile",
                        low_profile = "Low Profile",
                        motile = "Motile",
                        euplanktonic = "Euplanktonic",
                        algivore = "Algivore",
                        detritivore = "Detritivore",
                        plantivore = "Plantivore",
                        predator = "Predator"))

# axis labels
eig_vals <- final_model$CCA$eig
var_exp <- eig_vals / sum(eig_vals) * 100
axis1_lab <- paste0("CAP1 (", round(var_exp[1], 1), "% explained variance)")
axis2_lab <- paste0("CAP2 (", round(var_exp[2], 1), "% explained variance)")

global_rda <- ggplot() +
  geom_point(data = site_df, aes(CAP1, CAP2, color = age_ce), size = 4, alpha = 0.9) +
  geom_text_repel(data = species_df, aes(CAP1, CAP2, label = Label),
                  color = "red", size = 3.5, fontface = "italic", max.overlaps = Inf) +
  geom_segment(data = bp_df,
               aes(x = 0, y = 0, xend = CAP1, yend = CAP2),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black", linewidth = 0.7) +
  geom_text_repel(data = bp_df,
                  aes(CAP1, CAP2, label = Label),
                  size = 3.5, color = "black", max.overlaps = Inf) +
  scale_color_viridis_c(option = "D", name = "Age (CE)") +
  labs(x = axis1_lab, y = axis2_lab, title = NULL) +
  #coord_fixed() +
  theme_minimal(base_size = 14) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 1))

#global_rda

```

# Phase varpart 
## Phase varpart - Historical phases
```{r fig.width=10, fig.asp=0.5}

# ----------------------------------------------------------
# Phase varpart - Historical phases
# ----------------------------------------------------------

# --- Define predictor variables ---
selected_variables <- c("estimate", "NAO_Median_Value")

# --- Define phase subsets ---
phase_list <- list(
  "Phase 1" = joined_df_30yr %>% filter(age_ce < 750),
  "Phase 2" = joined_df_30yr %>% filter(age_ce >= 750 & age_ce < 1050),
  "Phase 3" = joined_df_30yr %>% filter(age_ce >= 1050 & age_ce < 1450),
  "Phase 4" = joined_df_30yr %>% filter(age_ce >= 1450 & age_ce < 1750),
  "Phase 5" = joined_df_30yr %>% filter(age_ce >= 1750)
)

# --- Function to compute pure and shared effects ---
#   implemented in `functions/source_custom_functions.R` (calculate_pure_shared_phase)

# --- Run across all phases ---
r2_partitioned <- imap_dfr(phase_list, calculate_pure_shared_phase)

# --- Add readable and ordered phase labels ---
r2_partitioned <- r2_partitioned %>%
  mutate(
    phase_label = factor(case_when(
      phase == "Phase 1" ~ "Phase 1\n(<750 CE)",
      phase == "Phase 2" ~ "Phase 2\n(750–1050 CE)",
      phase == "Phase 3" ~ "Phase 3\n(1050–1450 CE)",
      phase == "Phase 4" ~ "Phase 4\n(1450–1750 CE)",
      phase == "Phase 5" ~ "Phase 5\n(>1750 CE)"
    ), levels = c(
      "Phase 1\n(<750 CE)",
      "Phase 2\n(750–1050 CE)",
      "Phase 3\n(1050–1450 CE)",
      "Phase 4\n(1450–1750 CE)",
      "Phase 5\n(>1750 CE)"
    ))
  )

# --- Add readable and ordered phase labels ---
r2_partitioned <- r2_partitioned %>%
  mutate(
    phase_label = factor(case_when(
      phase == "Phase 1" ~ "Phase 1\n(<750 CE)",
      phase == "Phase 2" ~ "Phase 2\n(750–1050 CE)",
      phase == "Phase 3" ~ "Phase 3\n(1050–1450 CE)",
      phase == "Phase 4" ~ "Phase 4\n(1450–1750 CE)",
      phase == "Phase 5" ~ "Phase 5\n(>1750 CE)"
    ), levels = c(
      "Phase 1\n(<750 CE)",
      "Phase 2\n(750–1050 CE)",
      "Phase 3\n(1050–1450 CE)",
      "Phase 4\n(1450–1750 CE)",
      "Phase 5\n(>1750 CE)"
    ))
  )

# --- Define muted color palette for components ---
component_colors <- c(
  "Pure Vegetation" = "#97D8C4",
  "Pure Climate" = "#F4B942",
  "Shared" = "#58B0E6"
)

# --- Plot variance partitioning per phase ---
original_varpart <- ggplot(r2_partitioned, aes(x = phase_label, y = value, fill = component)) +
  geom_col(width = 0.6) +
  geom_text(
    data = r2_partitioned %>% filter(value > 0),
    aes(label = paste0(round(value * 100, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3.2
  ) +
  scale_fill_manual(values = component_colors, name = "Variance Component") +
  labs(
    x = NULL,
    y = "Adjusted R² (Variance explained)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(face = "bold"),
    strip.text.y.left = element_text(angle = 90, face = "bold"),
    plot.margin = ggplot2::margin(0.2, 0.2, 0.2, 0.2, unit = "lines")
  )

original_varpart

```
## Phase varpart - Historical phases - with significances
```{r fig.width=10, fig.asp=0.5}
# ============================================================
# PHASE varpart (YOUR PIPELINE) + SIGNIFICANCE + FADED (NS) FRACTIONS
# ============================================================

selected_variables <- c("NAO_Median_Value", "estimate")

phase_list <- list(
  "Phase 1" = joined_df_30yr %>% filter(age_ce < 750),
  "Phase 2" = joined_df_30yr %>% filter(age_ce >= 750 & age_ce < 1050),
  "Phase 3" = joined_df_30yr %>% filter(age_ce >= 1050 & age_ce < 1450),
  "Phase 4" = joined_df_30yr %>% filter(age_ce >= 1450 & age_ce < 1750),
  "Phase 5" = joined_df_30yr %>% filter(age_ce >= 1750)
)

# calculate_pure_shared_phase() in functions/source_custom_functions.R
r2_partitioned <- imap_dfr(phase_list, calculate_pure_shared_phase)

r2_partitioned <- r2_partitioned %>%
  mutate(
    phase_label = factor(case_when(
      phase == "Phase 1" ~ "Phase 1\n(<750 CE)",
      phase == "Phase 2" ~ "Phase 2\n(750–1050 CE)",
      phase == "Phase 3" ~ "Phase 3\n(1050–1450 CE)",
      phase == "Phase 4" ~ "Phase 4\n(1450–1750 CE)",
      phase == "Phase 5" ~ "Phase 5\n(>1750 CE)"
    ), levels = c(
      "Phase 1\n(<750 CE)",
      "Phase 2\n(750–1050 CE)",
      "Phase 3\n(1050–1450 CE)",
      "Phase 4\n(1450–1750 CE)",
      "Phase 5\n(>1750 CE)"
    ))
  )

component_colors <- c(
  "Pure Vegetation" = "#97D8C4",
  "Pure Climate" = "#F4B942",
  "Shared" = "#58B0E6"
)

# ============================================================
# SIGNIFICANCE + FADED (NS) FRACTIONS
# ============================================================

guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")

A <- "NAO_Median_Value"
B <- "estimate"

n_perm <- 9999   # use 999 for quick test
min_n  <- 5

set.seed(1)

# safe_anova_p() is implemented in `functions/source_custom_functions.R`

phase_sig <- imap_dfr(phase_list, function(dat, phase_name) {

  sub <- dat %>%
    select(age_ce, n_lakes, all_of(guild_cols), all_of(c(A, B))) %>%
    filter(if_all(all_of(c(guild_cols, A, B)), ~ !is.na(.x)))

  n_bins <- nrow(sub)
  mean_n_lakes <- mean(sub$n_lakes, na.rm = TRUE)

  if (n_bins < min_n) {
    return(tibble(
      phase = phase_name,
      n_bins = n_bins,
      mean_n_lakes = mean_n_lakes,
      p_full = NA_real_,
      p_pure_A = NA_real_,
      p_pure_B = NA_real_,
      p_seq_A_then_B = NA_real_,
      p_seq_B_then_A = NA_real_
    ))
  }

  comm <- sub %>% select(all_of(guild_cols)) %>% as.matrix()
  env  <- sub %>% select(all_of(c(A, B))) %>% as.data.frame()

  # Full model
  mod_full <- vegan::rda(as.formula(paste("comm ~", paste(c(A, B), collapse = " + "))), data = env)
  p_full <- safe_anova_p(mod_full, permutations = n_perm)

  # Pure (unique) effects
  mod_pure_A <- vegan::rda(as.formula(paste("comm ~", A, "+ Condition(", B, ")")), data = env)
  mod_pure_B <- vegan::rda(as.formula(paste("comm ~", B, "+ Condition(", A, ")")), data = env)
  p_pure_A <- safe_anova_p(mod_pure_A, permutations = n_perm)
  p_pure_B <- safe_anova_p(mod_pure_B, permutations = n_perm)

  # Sequential “second term” tests (order-dependent; by="term")
  mod_A_then_B <- vegan::rda(as.formula(paste("comm ~", A, "+", B)), data = env)
  p_seq_A_then_B <- safe_anova_p(mod_A_then_B, permutations = n_perm, by = "term", which_row = 2)

  mod_B_then_A <- vegan::rda(as.formula(paste("comm ~", B, "+", A)), data = env)
  p_seq_B_then_A <- safe_anova_p(mod_B_then_A, permutations = n_perm, by = "term", which_row = 2)

  tibble(
    phase = phase_name,
    n_bins = n_bins,
    mean_n_lakes = mean_n_lakes,
    p_full = p_full,
    p_pure_A = p_pure_A,
    p_pure_B = p_pure_B,
    p_seq_A_then_B = p_seq_A_then_B,
    p_seq_B_then_A = p_seq_B_then_A
  )
})

# --- SI-ready table (effect sizes + p-values)
phase_effects_wide <- r2_partitioned %>%
  mutate(component = recode(component,
                            "Pure Climate" = paste0("Pure ", A),
                            "Pure Vegetation" = paste0("Pure ", B))) %>%
  select(phase, phase_label, component, value) %>%
  pivot_wider(names_from = component, values_from = value)

phase_results_table <- phase_sig %>%
  left_join(phase_effects_wide, by = "phase") %>%
  select(
    phase, n_bins, mean_n_lakes,
    `Pure NAO_Median_Value`, `Pure estimate`, Shared,
    p_full, p_pure_A, p_pure_B,
    p_seq_A_then_B, p_seq_B_then_A
  ) %>%
  arrange(phase)

phase_results_table

# ------------------------------------------------------------
# FADED BARS IF NON-SIGNIFICANT (like moving window)
# ------------------------------------------------------------

alpha_sig <- 1.00
alpha_ns  <- 0.25

# Shared opacity rule:
#   "always"     -> Shared always opaque
#   "full_model" -> Shared fades if full model is not significant
shared_rule <- "always"   # change to "full_model" if you prefer

phase_flags <- phase_sig %>%
  transmute(
    phase,
    sig_pure_A = p_pure_A < 0.05,
    sig_pure_B = p_pure_B < 0.05,
    sig_full   = p_full   < 0.05
  )

r2_partitioned_sig <- r2_partitioned %>%
  left_join(phase_flags, by = "phase") %>%
  mutate(
    alpha_flag = case_when(
      component == "Pure Climate"    ~ ifelse(sig_pure_A, alpha_sig, alpha_ns),
      component == "Pure Vegetation" ~ ifelse(sig_pure_B, alpha_sig, alpha_ns),
      component == "Shared" & shared_rule == "full_model" ~ ifelse(sig_full, alpha_sig, alpha_ns),
      component == "Shared" & shared_rule == "always"     ~ alpha_sig,
      TRUE ~ alpha_sig
    )
  )

historical_phase_varpart_sig <- ggplot(r2_partitioned_sig,
                            aes(x = phase_label, y = value, fill = component, alpha = alpha_flag)) +
  geom_col(width = 0.6) +
  geom_text(
    data = r2_partitioned_sig %>% filter(value > 0),
    aes(label = paste0(round(value * 100, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3.2,
    inherit.aes = TRUE
  ) +
  scale_alpha_identity() +   # use alpha_flag directly; no alpha legend
  scale_fill_manual(values = component_colors, name = "Variance Component") +
  labs(
    x = NULL,
    y = "Adjusted R² (Variance explained)",
    caption = ""
  ) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(face = "bold"),
    plot.margin = ggplot2::margin(0.2, 0.2, 0.2, 0.2, unit = "lines")
  )

historical_phase_varpart_sig
```
## Moving window varpart
```{r fig.width = 10, fig.asp = 0.5}
# ============================================================
# MOVING WINDOW varpart (YOUR PIPELINE) + SIGNIFICANCE EXPLORATION
# + TRANSPARENCY WHEN NOT SIGNIFICANT (FINAL, FULL PIPELINE)
# ============================================================

selected_variables <- c("NAO_Median_Value", "estimate")

window_size <- 300
step_size   <- 30

min_year <- floor(min(joined_df_30yr$age_ce, na.rm = TRUE))
max_year <- ceiling(max(joined_df_30yr$age_ce, na.rm = TRUE))
window_starts <- seq(min_year, max_year - window_size, by = step_size)

r2_windowed <- map_dfr(
  window_starts,
  ~calculate_pure_shared_window(joined_df_30yr, .x, window_size, selected_variables)
) %>%
  mutate(midpoint = (window_start + window_end) / 2)

# Baseline plot (as before)
varpart_moving <- ggplot() +
  geom_col(
    data = r2_windowed,
    aes(x = midpoint, y = value, fill = component),
    width = step_size
  ) +
  scale_fill_manual(values = component_colors, name = "Variance Component") +
  labs(x = "Year (CE)", y = "Adjusted R² (Variance explained)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right", axis.text.y = element_text(face = "bold"))

varpart_moving
```

## Moving window varpart - with significance
```{r fig.width = 10, fig.asp = 0.5}
# ============================================================
# SIGNIFICANCE EXPLORATION (NEW, FIXED, ROBUST TO A/B NAMES)
# ============================================================

guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")

A <- "NAO_Median_Value"
B <- "estimate"

n_perm <- 9999   # use 999 for quick runs
min_n  <- 5

set.seed(1)

# calc_window_significance() in functions/source_custom_functions.R

# Run significance for all windows (same window_starts)
window_sig <- map_dfr(
  window_starts,
  ~calc_window_significance(
    joined_df_30yr,
    window_start = .x,
    window_size  = window_size,
    guild_cols   = guild_cols,
    A = A, B = B,
    min_n = min_n,
    n_perm = n_perm
  )
)

# ============================================================
# APPENDIX TABLE: merge effect sizes + p-values (one row/window)
# ============================================================

window_effects_wide <- r2_windowed %>%
  mutate(component = recode(component,
                            "Pure Climate" = paste0("Pure ", A),
                            "Pure Vegetation" = paste0("Pure ", B))) %>%
  select(window_start, window_end, midpoint, component, value) %>%
  pivot_wider(names_from = component, values_from = value)

window_results_table <- window_sig %>%
  left_join(window_effects_wide, by = c("window_start","window_end","midpoint")) %>%
  select(
    window_start, window_end, midpoint,
    n_bins, mean_n_lakes,
    !!paste0("Pure ", A), !!paste0("Pure ", B), Shared,
    p_full, p_pure_A, p_pure_B,
    p_seq_A_then_B, p_seq_B_then_A
  ) %>%
  arrange(midpoint)

window_results_table

# Add “stars” for convenience
window_results_table_stars <- window_results_table %>%
  mutate(
    full_sig = case_when(p_full < 0.001 ~ "***", p_full < 0.01 ~ "**", p_full < 0.05 ~ "*", TRUE ~ ""),
    pure_A_sig = case_when(p_pure_A < 0.001 ~ "***", p_pure_A < 0.01 ~ "**", p_pure_A < 0.05 ~ "*", TRUE ~ ""),
    pure_B_sig = case_when(p_pure_B < 0.001 ~ "***", p_pure_B < 0.01 ~ "**", p_pure_B < 0.05 ~ "*", TRUE ~ "")
  )

window_results_table_stars

summary_sig_counts <- window_results_table_stars %>%
  summarise(
    n_windows      = n(),
    n_full_sig     = sum(p_full < 0.05, na.rm = TRUE),
    n_pure_A_sig   = sum(p_pure_A < 0.05, na.rm = TRUE),
    n_pure_B_sig   = sum(p_pure_B < 0.05, na.rm = TRUE)
  )

summary_sig_counts


# ============================================================
# PLOT: fade bars when that component is NOT significant
# ------------------------------------------------------------
# Rules:
#   - Pure Climate opacity depends on p_pure_A
#   - Pure Vegetation opacity depends on p_pure_B
#   - Shared: you choose the rule below
# ============================================================

alpha_sig <- 1.00   # opaque when significant
alpha_ns  <- 0.25   # faded when NOT significant

# Shared opacity rule:
#   "always"    = always opaque
#   "full_model"= fade Shared when full model is not significant
shared_rule <- "always"  # change to "full_model" if you prefer

sig_flags <- window_sig %>%
  transmute(
    window_start, window_end, midpoint,
    sig_pure_A = p_pure_A < 0.05,
    sig_pure_B = p_pure_B < 0.05,
    sig_full   = p_full   < 0.05
  )

r2_windowed_sig <- r2_windowed %>%
  left_join(sig_flags, by = c("window_start","window_end","midpoint")) %>%
  mutate(
    alpha_flag = case_when(
      component == "Pure Climate"     ~ ifelse(sig_pure_A, alpha_sig, alpha_ns),
      component == "Pure Vegetation"  ~ ifelse(sig_pure_B, alpha_sig, alpha_ns),
      component == "Shared" & shared_rule == "full_model" ~ ifelse(sig_full, alpha_sig, alpha_ns),
      component == "Shared" & shared_rule == "always"     ~ alpha_sig,
      TRUE ~ alpha_sig
    )
  )

# ---- Define common x-scale ----
x_min <- floor(min(joined_df_30yr$age_ce, na.rm = TRUE) / 100) * 100
x_max <- ceiling(max(joined_df_30yr$age_ce, na.rm = TRUE) / 100) * 100

common_breaks  <- seq(x_min, x_max, by = 200)   # major grid
common_minor   <- seq(x_min, x_max, by = 100)   # minor grid

moving_window_varpart_sig <- ggplot() +
  geom_col(
    data = r2_windowed_sig,
    aes(x = midpoint, y = value, fill = component, alpha = alpha_flag),
    width = step_size
  ) +
  scale_alpha_identity() +  # use alpha_flag directly; no alpha legend
  scale_fill_manual(values = component_colors, name = "Variance Component") +
  scale_x_continuous(
   limits = c(x_min, x_max),
   breaks = common_breaks,
   minor_breaks = common_minor,
   expand = c(0, 0)
   ) +
  labs(
    x = "Year (CE)",
    y = "Adjusted R² (Variance explained)",
    caption = ""
  ) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "right", axis.text.y = element_text(face = "bold"))

moving_window_varpart_sig
```

## Effect sizes with significances
```{r}
# ----------------------------------------------------------
# Moving window varpart - Effect sizes (WORKING VERSION)
# ----------------------------------------------------------

# --- Predictor variables ---
selected_variables <- c("NAO_Median_Value", "estimate")

# --- Moving window parameters ---
window_size <- 300   # years
step_size   <- 30    # sliding step in years

# --- Define window sequence ---
min_year <- floor(min(joined_df_30yr$age_ce, na.rm = TRUE))
max_year <- ceiling(max(joined_df_30yr$age_ce, na.rm = TRUE))
window_starts <- seq(min_year, max_year - window_size, by = step_size)

# --- Run moving window analysis ---
# expects calculate_pure_shared_window(joined_df_30yr, window_start, window_size, selected_variables)
# and returns: window_start, window_end, component, value
r2_windowed <- map_dfr(
  window_starts,
  ~calculate_pure_shared_window(joined_df_30yr, .x, window_size, selected_variables)
) %>%
  mutate(midpoint = (window_start + window_end) / 2)

# ----------------------------------------------------------
# Effect size difference (Veg - Climate) + join significance
# ----------------------------------------------------------
# window_sig must exist and contain: window_start, window_end, midpoint, p_pure_A, p_pure_B
# IMPORTANT: set which p-value corresponds to which fraction:
# Here I assume:
#   p_pure_A = Pure Climate
#   p_pure_B = Pure Vegetation
# If your function returns the opposite, just swap them below.

stopifnot(exists("window_sig"))

effect_diff <- r2_windowed %>%
  filter(component %in% c("Pure Vegetation", "Pure Climate")) %>%
  select(window_start, window_end, midpoint, component, value) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  left_join(
    window_sig %>% select(window_start, window_end, midpoint, p_pure_A, p_pure_B),
    by = c("window_start", "window_end", "midpoint")
  ) %>%
  mutate(
    effect_diff = `Pure Vegetation` - `Pure Climate`,
    effect_group = case_when(
      effect_diff > 0 & midpoint < 750  ~ "VegChange > Climate (pre-750)",
      effect_diff > 0 & midpoint >= 750 ~ "VegChange > Climate (post-750)",
      effect_diff < 0                   ~ "Climate > VegChange",
      TRUE                              ~ NA_character_
    ),
    sig_clim = !is.na(p_pure_A) & p_pure_A < 0.05,
    sig_veg  = !is.na(p_pure_B) & p_pure_B < 0.05,
    direction = case_when(
      effect_diff > 0 ~ "veg",
      effect_diff < 0 ~ "clim",
      TRUE            ~ NA_character_
    ),
    alpha_flag = case_when(
      direction == "veg"  & sig_veg  ~ 1.0,
      direction == "veg"  & !sig_veg ~ 0.25,
      direction == "clim" & sig_clim ~ 1.0,
      direction == "clim" & !sig_clim~ 0.25,
      TRUE ~ 0.25
    )
  )

effect_diff <- effect_diff %>%
  filter(!is.na(effect_group))

# --- Colors ---
effect_colors_split <- c(
  "VegChange > Climate (pre-750)"  = "#054A29",
  "VegChange > Climate (post-750)" = "#6DC6B6",
  "Climate > VegChange"            = "#F4B942"
)

# If you don't have phase_lines defined, comment this out or define it:
# phase_lines <- c(1000, 1450, 1750)   # example
if (!exists("phase_lines")) phase_lines <- numeric(0)


# ---- Define common x-scale ----
x_min <- floor(min(joined_df_30yr$age_ce, na.rm = TRUE) / 100) * 100
x_max <- ceiling(max(joined_df_30yr$age_ce, na.rm = TRUE) / 100) * 100

common_breaks  <- seq(x_min, x_max, by = 200)   # major grid
common_minor   <- seq(x_min, x_max, by = 100)   # minor grid

# --- Plot ---
effect_diff_plot_sig <- ggplot(
  effect_diff,
  aes(x = midpoint, y = effect_diff, fill = effect_group, alpha = alpha_flag)
) +
  geom_col(width = step_size) +
  scale_fill_manual(
  values = effect_colors_split,
  name = "Dominant Driver",
  drop = TRUE
  ) +
  scale_alpha_identity() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = phase_lines, linetype = "dotted", color = "gray30") +
  scale_x_continuous(
    limits = c(x_min, x_max),
    breaks = common_breaks,
    minor_breaks = common_minor,
    expand = c(0, 0)
    ) +
  labs(
    x = "Year (CE)",
    y = "Effect size (Vegetation − Climate)",
    title = "(c) Relative dominance of VegChange vs. climate",
    caption = ""
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = ggplot2::margin(0.2, 0.2, 0.2, 0.2, unit = "lines")
  )

effect_diff_plot_sig
```

# Sensitivty analyses
## Sensitivty analyses - PHASE-BOUNDARY ROBUSTNESS (GRID TEST)
```{r}
# ============================================================
# SENSITIVITY: PHASE-BOUNDARY ROBUSTNESS (GRID TEST)
# ============================================================

guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")

env_predictors <- c("NAO_Median_Value", "estimate")
A <- env_predictors[1]  # NAO
B <- env_predictors[2]  # Veg

# -------------------------
# calc_phase_varpart(), make_phase_breaks(), run_one_boundary_set()
# are in `functions/source_custom_functions.R`
# -------------------------

# ---- baseline boundaries ----
base_bounds <- c(750, 1050, 1450, 1750)

# ---- settings: shift boundaries by k bins (bin=30 yrs) ----
bin_size <- 30
max_shift_bins <- 2  # +/- 2 bins => +/- 60 years
shift_vals <- seq(-max_shift_bins, max_shift_bins) * bin_size

# ------------------------------------------------------------
# Grid of boundary perturbations
# - enforce ordering: b1 < b2 < b3 < b4
# - enforce minimum bins per phase (min_n)
# ------------------------------------------------------------
min_n <- 5  # same as your phase varpart threshold

boundary_grid <- tidyr::crossing(
  d1 = shift_vals,
  d2 = shift_vals,
  d3 = shift_vals,
  d4 = shift_vals
) %>%
  mutate(
    b1 = base_bounds[1] + d1,
    b2 = base_bounds[2] + d2,
    b3 = base_bounds[3] + d3,
    b4 = base_bounds[4] + d4
  ) %>%
  filter(b1 < b2, b2 < b3, b3 < b4) %>%
  mutate(grid_id = row_number())

# ------------------------------------------------------------
# Baseline run (for comparison)
# ------------------------------------------------------------
phase_breaks_base <- make_phase_breaks(base_bounds[1], base_bounds[2], base_bounds[3], base_bounds[4])

r2_base_phase <- calc_phase_varpart(
  joined_df_30yr,
  guild_cols = guild_cols,
  env_predictors = env_predictors,
  phase_breaks = phase_breaks_base,
  min_n = min_n
)

base_wide <- r2_base_phase %>%
  select(phase, component, value, n_bins) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  rename(base_n_bins = n_bins)

# ------------------------------------------------------------
# Run varpart for each perturbed boundary set
# ------------------------------------------------------------
r2_phase_boundary_sens_long <- pmap_dfr(
  boundary_grid %>% select(b1, b2, b3, b4, grid_id),
  run_one_boundary_set
)

# ------------------------------------------------------------
# Compare each run to baseline, phase-by-phase
# ------------------------------------------------------------
r2_phase_boundary_sens <- r2_phase_boundary_sens_long %>%
  left_join(
    r2_base_phase %>%
      select(phase, component, base_value = value, base_n_bins = n_bins),
    by = c("phase","component")
  ) %>%
  mutate(
    delta_value = value - base_value,
    abs_delta   = abs(delta_value)
  )

# ------------------------------------------------------------
# Summaries you can report in SI:
# - max absolute deviation from baseline by component x phase
# - how often sign/dominance flips (veg vs climate) by phase
# ------------------------------------------------------------
A <- env_predictors[1]  # NAO
B <- env_predictors[2]  # estimate

sens_summary_phase_component <- r2_phase_boundary_sens %>%
  group_by(phase, component) %>%
  summarise(
    n_runs = n(),
    max_abs_delta = max(abs_delta, na.rm = TRUE),
    median_abs_delta = median(abs_delta, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(phase, component)

# dominance flips: whether Pure B >= Pure A within a phase
sens_dom <- r2_phase_boundary_sens_long %>%
  filter(component %in% c(paste0("Pure ", A), paste0("Pure ", B))) %>%
  select(grid_id, phase, component, value) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  mutate(
    dom_B = .data[[paste0("Pure ", B)]] >= .data[[paste0("Pure ", A)]]
  ) %>%
  group_by(phase) %>%
  summarise(
    prop_runs_B_dominant = mean(dom_B, na.rm = TRUE),
    .groups = "drop"
  )

sens_summary_phase_component
sens_dom
```

## Sensitivty analyses - Leave-one-out per phases
```{r}
# -------------------------
# Settings
# -------------------------
guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")

env_predictors <- c("NAO_Median_Value", "estimate")
A <- env_predictors[1]  # NAO
B <- env_predictors[2]  # Veg

# Phase definitions (same as your plot)
phase_breaks <- list(
  "Phase 1" = c(-Inf, 750),
  "Phase 2" = c(750, 1050),
  "Phase 3" = c(1050, 1450),
  "Phase 4" = c(1450, 1750),
  "Phase 5" = c(1750, Inf)
)

# -------------------------
# Helper: rebuild joined table from a bio_reg_30 (for LOO)
# (consistent with your Step 5+6: align then scale)
# -------------------------
# build_joined_from_bio_reg() is implemented in `functions/source_custom_functions.R`

# -------------------------
# 1) BASELINE (all lakes) — phase varpart
# -------------------------
r2_base_phase <- calc_phase_varpart(
  joined_df_30yr,
  guild_cols = guild_cols,
  env_predictors = env_predictors,
  phase_breaks = phase_breaks,
  min_n = 5
)

# wide baseline for deltas
base_wide <- r2_base_phase %>%
  select(phase, component, value) %>%
  pivot_wider(names_from = component, values_from = value)

# overall baseline summaries (across phases)
base_summary <- r2_base_phase %>%
  filter(component %in% c(paste0("Pure ", A), paste0("Pure ", B), "Shared")) %>%
  group_by(component) %>%
  summarise(
    base_mean   = mean(value, na.rm = TRUE),
    base_median = median(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = component,
              values_from = c(base_mean, base_median))

# -------------------------
# 2) LOO — recompute bio_reg_30 then joined then phase varpart
# -------------------------
lakes <- sort(unique(bio_lake_30$lake))

r2_loo_phase <- map_dfr(lakes, function(lk) {

  bio_reg_loo <- bio_lake_30 %>%
    filter(lake != lk) %>%
    group_by(age_ce) %>%
    summarise(
      n_lakes = n_distinct(lake),
      across(all_of(guild_cols), ~ mean(.x, na.rm = TRUE)),
      .groups = "drop"
    )

  joined_loo <- build_joined_from_bio_reg(
    bio_reg_loo,
    nao_30 = nao_30, nhst_30 = nhst_30, veg_30 = veg_30, prop_comm_30 = prop_comm_30,
    env_cols = c("NAO_Median_Value","Rmean","estimate")
  )

  out <- calc_phase_varpart(
    joined_loo,
    guild_cols = guild_cols,
    env_predictors = env_predictors,
    phase_breaks = phase_breaks,
    min_n = 5
  )

  out %>% mutate(leave_out = lk)
})

# -------------------------
# Summary table
# -------------------------
loo_phase_wide <- r2_loo_phase %>%
  select(leave_out, phase, component, value) %>%
  pivot_wider(names_from = component, values_from = value)

# dominance per phase: B (veg) >= A (NAO)
loo_phase_dom <- loo_phase_wide %>%
  mutate(
    dominant = ifelse(.data[[paste0("Pure ", B)]] >= .data[[paste0("Pure ", A)]], B, A)
  ) %>%
  group_by(leave_out) %>%
  summarise(
    n_phases = n(),
    prop_phases_B_dominant = mean(dominant == B, na.rm = TRUE),
    prop_phases_A_dominant = mean(dominant == A, na.rm = TRUE),
    .groups = "drop"
  )

# effect sizes averaged across phases
loo_phase_effects <- loo_phase_wide %>%
  group_by(leave_out) %>%
  summarise(
    mean_pure_A    = mean(.data[[paste0("Pure ", A)]], na.rm = TRUE),
    mean_pure_B    = mean(.data[[paste0("Pure ", B)]], na.rm = TRUE),
    mean_shared    = mean(.data[["Shared"]], na.rm = TRUE),
    median_pure_A  = median(.data[[paste0("Pure ", A)]], na.rm = TRUE),
    median_pure_B  = median(.data[[paste0("Pure ", B)]], na.rm = TRUE),
    median_shared  = median(.data[["Shared"]], na.rm = TRUE),
    mean_diff_BA   = mean(.data[[paste0("Pure ", B)]] - .data[[paste0("Pure ", A)]], na.rm = TRUE),
    .groups = "drop"
  )

# deltas vs baseline (across phases)
base_means <- r2_base_phase %>%
  filter(component %in% c(paste0("Pure ", A), paste0("Pure ", B), "Shared")) %>%
  group_by(component) %>%
  summarise(base_mean = mean(value, na.rm = TRUE), .groups="drop") %>%
  deframe()

loo_phase_vs_base <- loo_phase_effects %>%
  mutate(
    delta_mean_pure_A = mean_pure_A - base_means[[paste0("Pure ", A)]],
    delta_mean_pure_B = mean_pure_B - base_means[[paste0("Pure ", B)]],
    delta_mean_shared = mean_shared - base_means[["Shared"]]
  )

# final appendix table
loo_phase_summary <- loo_phase_vs_base %>%
  left_join(loo_phase_dom, by = "leave_out") %>%
  arrange(desc(abs(delta_mean_pure_B)))  # or any ordering you like

loo_phase_summary

# OPTIONAL: per-phase values in wide form for appendix too
# (one row per leave-out, columns per phase x component)
loo_phase_perphase <- r2_loo_phase %>%
  mutate(phase_component = paste(phase, component, sep = " — ")) %>%
  select(leave_out, phase_component, value) %>%
  pivot_wider(names_from = phase_component, values_from = value)

loo_phase_perphase
```

## Sensitivity analyses - Leave one out - moving window
```{r}
# ------------------------------------------------------------
# Inputs that already exist from your pipeline above:
#   bio_lake_30      (lake-level binned biology)
#   joined_df_30yr   (regional biology + aligned env + scaled)
# ------------------------------------------------------------

guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")

# predictors (already scaled in joined_df_30yr)
env_predictors <- c("NAO_Median_Value", "estimate")  # add "Rmean" if you want later

# ------------------------------------------------------------
# Window varpart working DIRECTLY on joined_df_30yr
# calc_window_varpart_joined() in functions/source_custom_functions.R
# ------------------------------------------------------------

# run
guild_cols <- c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator")
env_predictors <- c("NAO_Median_Value", "estimate")

# Run
r2_base <- calc_window_varpart_joined(
  joined_df_30yr,
  window_size = 300, step_size = 30, min_bins = 5,
  guild_cols = c("high_profile","low_profile","motile","euplanktonic",
                "algivore","detritivore","plantivore","predator"),
  env_predictors = c("NAO_Median_Value", "estimate")
)

# Base result using the already-aligned joined_df_30yr
r2_base <- calc_window_varpart_joined(joined_df_30yr,
                                      window_size = 300, step_size = 30, min_bins = 5)

# ------------------------------------------------------------
# Leave-one-out: recompute bio_reg_30 (regional) from bio_lake_30,
# then rebuild joined table on SAME age_ce grid, then scale, then varpart
# build_joined_from_bio_reg() in functions/source_custom_functions.R
# ------------------------------------------------------------

lakes <- sort(unique(bio_lake_30$lake))

r2_loo <- map_dfr(lakes, function(lk) {

  bio_reg_loo <- bio_lake_30 %>%
    filter(lake != lk) %>%
    group_by(age_ce) %>%
    summarise(
      n_lakes = n_distinct(lake),
      across(all_of(guild_cols), ~mean(.x, na.rm = TRUE)),
      .groups = "drop"
    )

  joined_loo <- build_joined_from_bio_reg(
    bio_reg_loo,
    nao_30 = nao_30, nhst_30 = nhst_30, veg_30 = veg_30, prop_comm_30 = prop_comm_30
  )

  out <- calc_window_varpart_joined(joined_loo,
                                    window_size = 300, step_size = 30, min_bins = 5)

  out %>% mutate(leave_out = lk)
})

dominance_loo <- r2_loo %>%
  select(leave_out, midpoint, component, value) %>%
  tidyr::pivot_wider(names_from = component, values_from = value) %>%
  mutate(dominant = ifelse(.data[[paste0("Pure ", env_predictors[2])]] >=
                             .data[[paste0("Pure ", env_predictors[1])]],
                           "SecondPredictor", "FirstPredictor")) %>%
  group_by(leave_out) %>%
  summarise(prop_windows_second_pred_dominant = mean(dominant == "SecondPredictor", na.rm = TRUE),
            .groups = "drop")

A <- env_predictors[1]  # "NAO_Median_Value"
B <- env_predictors[2]  # "estimate"

# ---- 1) Summarise effect sizes per lake (mean/median across windows) ----
loo_effects <- r2_loo %>%
  filter(component %in% c(paste0("Pure ", A), paste0("Pure ", B))) %>%
  select(leave_out, midpoint, component, value) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  group_by(leave_out) %>%
  summarise(
    mean_pure_A    = mean(.data[[paste0("Pure ", A)]], na.rm = TRUE),
    mean_pure_B    = mean(.data[[paste0("Pure ", B)]], na.rm = TRUE),
    median_pure_A  = median(.data[[paste0("Pure ", A)]], na.rm = TRUE),
    median_pure_B  = median(.data[[paste0("Pure ", B)]], na.rm = TRUE),
    mean_diff_BA   = mean(.data[[paste0("Pure ", B)]] - .data[[paste0("Pure ", A)]], na.rm = TRUE),
    n_windows      = dplyr::n(),
    .groups = "drop"
  )

# ---- 2) Your existing dominance metric (keep it) ----
dominance_loo <- r2_loo %>%
  select(leave_out, midpoint, component, value) %>%
  tidyr::pivot_wider(names_from = component, values_from = value) %>%
  mutate(
    dominant = ifelse(
      .data[[paste0("Pure ", B)]] >= .data[[paste0("Pure ", A)]],
      B, A
    )
  ) %>%
  group_by(leave_out) %>%
  summarise(
    prop_windows_B_dominant = mean(dominant == B, na.rm = TRUE),
    prop_windows_A_dominant = mean(dominant == A, na.rm = TRUE),
    .groups = "drop"
  )

# ---- 3) Combine into one tidy table ----
loo_summary <- loo_effects %>%
  left_join(dominance_loo, by = "leave_out")

loo_summary

# baseline (all lakes)
base_effects <- r2_base %>%
  filter(component %in% c(paste0("Pure ", A), paste0("Pure ", B))) %>%
  select(midpoint, component, value) %>%
  pivot_wider(names_from = component, values_from = value) %>%
  summarise(
    base_mean_pure_A = mean(.data[[paste0("Pure ", A)]], na.rm = TRUE),
    base_mean_pure_B = mean(.data[[paste0("Pure ", B)]], na.rm = TRUE)
  )

loo_summary_vs_base <- loo_summary %>%
  mutate(
    delta_mean_pure_A = mean_pure_A - base_effects$base_mean_pure_A,
    delta_mean_pure_B = mean_pure_B - base_effects$base_mean_pure_B
  )

loo_summary_vs_base

```

## Sensitivity analyses - lake record position / length
```{r fig.width=12}
# ------------------------------------------------------------
# Null model: circularly shift AGE BINS within each lake on bio_lake_30,
# then rebuild regional + joined, then varpart
# ------------------------------------------------------------
set.seed(1)

# circular_shift_by_lake() is implemented in `functions/source_custom_functions.R`

n_null <- 500

r2_null <- map_dfr(seq_len(n_null), function(i) {

  bio_null <- circular_shift_by_lake(bio_lake_30)

  bio_reg_null <- bio_null %>%
    group_by(age_ce) %>%
    summarise(
      n_lakes = n_distinct(lake),
      across(all_of(guild_cols), ~mean(.x, na.rm = TRUE)),
      .groups = "drop"
    )

  joined_null <- build_joined_from_bio_reg(
    bio_reg_null,
    nao_30 = nao_30, nhst_30 = nhst_30, veg_30 = veg_30, prop_comm_30 = prop_comm_30
  )

  out <- calc_window_varpart_joined(joined_null,
                                    window_size = 300, step_size = 30, min_bins = 5)

  out %>% mutate(null_iter = i)
})

null_env <- r2_null %>%
  group_by(midpoint, component) %>%
  summarise(
    q05 = quantile(value, 0.05, na.rm = TRUE),
    q50 = quantile(value, 0.50, na.rm = TRUE),
    q95 = quantile(value, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# Panel 1: Pure Vegetation = Pure estimate
obs_veg <- r2_base %>% dplyr::filter(component == "Pure estimate")
nul_veg <- null_env %>% dplyr::filter(component == "Pure estimate")

p_veg <- ggplot() +
 #geom_ribbon(
 #  data = nul_veg,
 #  aes(x = midpoint, ymin = q05, ymax = q95),
 #  fill = "grey80", alpha = 0.5
 #) +
  geom_line(
    data = nul_veg,
    aes(x = midpoint, y = q50),
    linewidth = 0.7, color = "grey30"
  ) +
  geom_line(
    data = obs_veg,
    aes(x = midpoint, y = value),
    linewidth = 1.1, color = "black" 
  ) +
  labs(
    x = "Year (CE)",
    y = "Adjusted R² (Pure Vegetation)",
    title = "Pure Vegetation: observed vs null envelope"
  ) +
  theme_minimal(base_size = 13)

# Panel 2: Pure Climate = Pure NAO_Median_Value
obs_clim <- r2_base %>% dplyr::filter(component == "Pure NAO_Median_Value")
nul_clim <- null_env %>% dplyr::filter(component == "Pure NAO_Median_Value")

p_clim <- ggplot() +
  #geom_ribbon(
  #  data = nul_clim,
  #  aes(x = midpoint, ymin = q05, ymax = q95),
  #  fill = "grey80", alpha = 0.5
  #) +
  geom_line(
    data = nul_clim,
    aes(x = midpoint, y = q50),
    linewidth = 0.7, color = "grey30"
  ) +
  geom_line(
    data = obs_clim,
    aes(x = midpoint, y = value),
    linewidth = 1.1, color = "black"
  ) +
  labs(
    x = "Year (CE)",
    y = "Adjusted R² (Pure Climate)",
    title = "Pure Climate: observed vs null envelope"
  ) +
  theme_minimal(base_size = 13)

# plot_obs_vs_null(), plot_obs_vs_null_points() in functions/source_custom_functions.R

p1 <- plot_obs_vs_null("Pure estimate")
p2 <- plot_obs_vs_null("Pure NAO_Median_Value")

p_veg <- plot_obs_vs_null_points(
  component_name = "Pure estimate",
  ylab = "Adjusted R² (Pure Vegetation)",
  title_text = "Pure Vegetation"
)

p_clim <- plot_obs_vs_null_points(
  component_name = "Pure NAO_Median_Value",
  ylab = "Adjusted R² (Pure Climate)",
  title_text = "Pure Climate (NAO)"
)

p_veg  <- p_veg  + coord_cartesian(ylim = c(0, 0.6))
p_clim <- p_clim + coord_cartesian(ylim = c(0, 0.6))

combined_plot <- p_veg / p_clim +
  plot_annotation(
    title = "Observed vs null median for vegetation and climate effects",
    theme = theme(plot.title = element_text(face = "bold"))
  )

#combined_plot

```

# FIGURES
### FIGURE 2a: NAO panel on top + main panel (DCA + vegetation)
```{r fig.asp=0.75, fig.width=12}
# ----------------------------------------------------------
# Combine smoothed GAM outputs and DCA scores across groups
# ----------------------------------------------------------

smooth_all_in_one <- 
  bind_rows(
    cbind(type = rep("diatoms", nrow(smooth_diatoms)), smooth_diatoms),
    cbind(type = rep("chiro", nrow(smooth_diatoms)), smooth_chiro)
  )

df_all_in_one <- 
  bind_rows(
    cbind(type = rep("diatoms", nrow(b_ls_dca_diatoms)), b_ls_dca_diatoms),
    cbind(type = rep("chiro", nrow(b_ls_dca_chiro)), b_ls_dca_chiro)
  )

# ==========================================================
# Fig 1b (FINAL): NAO panel on top + main panel (DCA + vegetation)
# ==========================================================

# -----------------------------
# 1) Species smooth (diatoms + chiro)
# -----------------------------
smooth_main <- smooth_all_in_one %>%
  filter(.smooth == "s(age_ce)") %>%
  mutate(type = factor(type, levels = c("diatoms", "chiro")))

smooth_lines <- smooth_main %>%
  mutate(group_label = if_else(type == "diatoms", "Producers", "Consumers"))

# -----------------------------
# 2) Vegetation smooth (from sm_to)
# -----------------------------
veg_bg <- sm_to %>%
  filter(.smooth == "s(age_ce)") %>%
  transmute(
    age_ce  = age_ce,
    veg_est = .estimate,
    veg_lo  = .lower_ci,
    veg_hi  = .upper_ci
  ) %>%
  arrange(age_ce)

# -----------------------------
# 3) Map vegetation onto DCA axis (fixed veg range)
# -----------------------------
veg_limits <- c(-1.5, 1.5)
dca_rng <- range(c(smooth_main$.lower_ci, smooth_main$.upper_ci), na.rm = TRUE)

# veg_to_dca() and dca_to_veg() are implemented in `functions/source_custom_functions.R`

# -----------------------------
# 4) NAO series (median per age_ce, centered anomaly)
# -----------------------------
nao_series <- nao_hernandez %>%
  group_by(age_ce) %>%
  summarise(nao = median(NAO_Median_Value, na.rm = TRUE), .groups = "drop") %>%
  arrange(age_ce) %>%
  filter(is.finite(nao)) %>%
  mutate(
    nao_anom = nao - median(nao, na.rm = TRUE),
    y0 = 0,
    y_pos = pmax(nao_anom, 0),
    y_neg = pmin(nao_anom, 0)
  )

# -----------------------------
# 5) Aesthetics
# -----------------------------
alpha_ci_species <- 0.10
alpha_ci_veg     <- 0.10
alpha_nao        <- 0.35

lw_species <- 3.6
lw_veg     <- 2.8

# Producers yellow, Consumers purple, Vegetation mid-viridis green
cols_line <- c(
  "Consumers"  = viridisLite::viridis(9)[1],
  "Vegetation" = viridisLite::viridis(9)[5],
  "Producers"  = viridisLite::viridis(9)[9]
)
cols_fill_species <- c(
  "Producers" = cols_line["Producers"],
  "Consumers" = cols_line["Consumers"]
)

# -----------------------------
# 6) TOP panel: NAO with normal axis (no x-axis shown)
# -----------------------------
p_nao <- ggplot(nao_series, aes(x = age_ce)) +
  geom_ribbon(aes(ymin = y0, ymax = y_pos), fill = "red",  alpha = alpha_nao) +
  geom_ribbon(aes(ymin = y_neg, ymax = y0), fill = "blue", alpha = alpha_nao) +
  geom_hline(yintercept = 0, linewidth = 0.4, alpha = 0.7) +
  scale_x_continuous(limits = c(0, 2020), n.breaks = 5) +
  labs(x = NULL, y = "NAO (median anomaly)") +
  theme_bw(base_size = 12) +
  theme(
    axis.title.x  = element_blank(),
    axis.text.x   = element_blank(),
    axis.ticks.x  = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = ggplot2::margin(5.5, 5.5, 0, 5.5)  # bottom margin = 0
  )

# -----------------------------
# 7) BOTTOM panel: DCA smooths + vegetation (sec.axis)
# -----------------------------
p_main <- ggplot() +
  # Vegetation (CI + line)
  geom_ribbon(
    data = veg_bg,
    aes(x = age_ce, ymin = veg_to_dca(veg_lo), ymax = veg_to_dca(veg_hi)),
    inherit.aes = FALSE,
    fill = cols_line["Vegetation"],
    alpha = alpha_ci_veg,
    show.legend = FALSE
  ) +
  geom_line(
    data = veg_bg,
    aes(x = age_ce, y = veg_to_dca(veg_est), colour = "Vegetation"),
    inherit.aes = FALSE,
    linewidth = lw_veg
  ) +
  # Species CI + lines
  geom_ribbon(
    data = smooth_lines,
    aes(x = age_ce, ymin = .lower_ci, ymax = .upper_ci, fill = group_label),
    alpha = alpha_ci_species,
    linewidth = 0,
    show.legend = FALSE
  ) +
  geom_line(
    data = smooth_lines,
    aes(x = age_ce, y = .estimate, colour = group_label),
    linewidth = lw_species
  ) +
  scale_colour_manual(
    name = NULL,
    values = cols_line,
    breaks = c("Producers", "Consumers", "Vegetation")
  ) +
  scale_fill_manual(values = cols_fill_species, guide = "none") +
  scale_y_continuous(
    name = "DCA1 (regional smooth)",
    limits = c(-1.75, 1),  # your final y-lims
    sec.axis = sec_axis(~ dca_to_veg(.), name = "Regional vegetation change (HGAM smooth)")
  ) +
  scale_x_continuous(limits = c(0, 2020), n.breaks = 5) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.margin = ggplot2::margin(0, 5.5, 5.5, 5.5)  # top margin = 0
  )

# -----------------------------
# 8) Stack with ZERO gap
# -----------------------------
final_fig <- p_nao / p_main +
  plot_layout(heights = c(0.55, 2)) &
  theme(panel.spacing = unit(0, "pt"))

final_fig

# -----------------------------
# Export
# -----------------------------
ggsave(
  "figures/FIGURE_2a_all_regional_smooths_with_NAO_panel.pdf",
  plot = final_fig,
  dpi = 300, width = 300, height = 150, units = "mm"
)
```

### FIGURE 2b - Individual lake DCAs plus map
```{r fig.asp=2, fig.width=12}
# ==========================================================
# FINAL FIGURE (NO PANEL A):
# Map of Azores + embedded lake DCA smooths (diatoms vs chiro)
# ==========================================================

# -----------------------------
# 1) Plot titles with island abbreviations
# -----------------------------
custom_titles <- c(
  "Azul" = "Azul [SM]",
  "Caldeirao" = "Caldeirão [CR]",
  "Caveiro" = "Caveiro [PI]",
  "Empadadas Norte" = "Empada. Norte [SM]",
  "Funda" = "Funda [FL]",
  "Ginjal" = "Ginjal [TE]",
  "Peixinho" = "Peixinho [PI]",
  "Prata" = "Prata [SM]",
  "Santiago" = "Santiago [SM]"
)

# -----------------------------
# 2) Build DCA mini-plots (diatoms vs chironomids) -> grobs
# -----------------------------
ls_plot_df_dca_multi_all <-
  ls_dca_multi %>%
  map(~ .x %>% filter(group %in% 1:2)) %>%
  imap(~ ggplot(.x, aes(x = age_ce, y = DCA1)) +
         geom_smooth(aes(colour = factor(group)),
                     linewidth = 1.4, alpha = 0.6, method = "gam", se = TRUE) +
         scale_colour_viridis_d(
           name = NULL,
           labels = c("Producers", "Consumers")
         ) +
         labs(title = .y, x = NULL, y = NULL) +
         theme_bw(base_size = 7) +
         theme(
           legend.position = "none",
           plot.title = element_text(face = "bold", size = 7, hjust = 0.5),
           axis.text = element_text(size = 6),
           panel.grid.minor = element_blank(),
           plot.margin = ggplot2::margin(1, 1, 1, 1)
         )
  )

ls_plot_df_dca_multi_all_named <- imap(ls_plot_df_dca_multi_all, function(p, nm) {
  if (nm %in% names(custom_titles)) p + ggtitle(custom_titles[[nm]]) else p
})

# remove lakes you don't want
ls_plot_df_dca_multi_all_named <- ls_plot_df_dca_multi_all_named[!names(ls_plot_df_dca_multi_all_named) %in% c("Furnas","Fogo")]

scatterplots <- map(ls_plot_df_dca_multi_all_named, ~ ggplotGrob(.x))

# -----------------------------
# 3) Lake points (sf)
# -----------------------------
lake_centers <- data.frame(
  lake = lake_metadata$lake,
  lat  = lake_metadata$lat,
  lon  = lake_metadata$long
)

island_points <- st_as_sf(lake_centers, coords = c("lon", "lat"), crs = 4326)

# -----------------------------
# 4) Bounding box for Azores map (adjust if needed)
# -----------------------------
xmin <- -32.0 - 0.75 * abs(-32.0 + 24.5)
xmax <- -24.5 + 0.75 * abs(-32.0 + 24.5)
ymin <-  36.5 - 0.5  * abs(40.0 - 36.5)
ymax <-  40.0 + 0.5  * abs(40.0 - 36.5)

azores_bbox <- st_as_sfc(st_bbox(c(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), crs = 4326))

# -----------------------------
# 5) Coastline + crop
# -----------------------------
coastline_shapefile <- st_read("data/ne_10m_coastline/ne_10m_coastline.shp", quiet = TRUE)
azores_coastline <- st_crop(coastline_shapefile, azores_bbox)

# -----------------------------
# 6) Where to place the embedded mini-plots
#     (must match the order of scatterplots)
# -----------------------------
scatterplot_positions <- list(
  c(xmax - 1.5, ymax - 1),                # Azul
  c(xmin + 1.5, ymax - 1),                # Caldeirão
  c(xmin + 1.5, ymin + 1),                # Caveiro
  c(xmax - 1.5, (ymax + ymin) / 2),       # Empadadas Norte
  c(xmin + 1.5, (ymax + ymin) / 2),       # Funda
  c(xmin + (xmax - xmin) / 2, ymin + 1),  # Ginjal
  c(((xmin + 1.5)+(xmin + (xmax - xmin) / 2))/2, ymin + 1),     # Peixinho
  c(((xmin + (xmax - xmin) / 2)+(xmax - 1.5))/2, ymin + 1),     # Prata
  c(xmax - 1.5, ymin + 1)                 # Santiago
)

# segments from inset positions to lake points
segment_data <- data.frame(
  x    = sapply(scatterplot_positions, `[[`, 1),
  y    = sapply(scatterplot_positions, `[[`, 2),
  xend = st_coordinates(island_points$geometry)[, 1],
  yend = st_coordinates(island_points$geometry)[, 2]
)

# -----------------------------
# 7) Build the final plot (single panel)
# -----------------------------
final_fig <- ggplot() +
  geom_sf(data = azores_coastline, fill = "lightblue", colour = "black", linewidth = 0.3) +
  geom_sf(data = island_points, shape = 21, size = 3.2, stroke = 0.3,
          fill = "grey70", colour = "white") +
  geom_segment(
    data = segment_data,
    aes(x = x, y = y, xend = xend, yend = yend),
    colour = "grey60", linewidth = 0.3
  ) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), datum = sf::st_crs(4326), expand = FALSE) +
  theme_void() +
  theme(
    plot.margin = ggplot2::margin(1, 1, 1, 1)  # IMPORTANT: no outer whitespace
  ) +
  ggtitle("(b) Detailed map of the Azores Archipelago") +
  theme(plot.title = element_text(size = 9, face = "bold", hjust = -2)) +
  coord_sf(xlim = c(xmin, xmax + 1.0), ylim = c(ymin, ymax), expand = FALSE, clip = "off") +
  theme(plot.margin = ggplot2::margin(0, 20, 0, 20))   # add right margin (pt)

final_fig <- final_fig +
  labs(x = "Year (CE)") +
  theme(
    axis.title.x = element_text(size = 10, margin = ggplot2::margin(t = 6)),
    axis.text.x  = element_text(size = 8),
    axis.ticks.x = element_line(linewidth = 0.3),
    #axis.line.x  = element_line(linewidth = 0.3)
  )

# Optional north arrow (remove if not needed)
final_fig <- final_fig +
  annotation_north_arrow(
    location = "bl", which_north = "true",
    pad_x = unit(0.25, "in"), pad_y = unit(0.25, "in"),
    style = north_arrow_minimal()
  )

# -----------------------------
# 8) Overlay the inset plots
# -----------------------------
scatterplot_scale_x <- 2 * 0.85
scatterplot_scale_y <- 1.5 * 0.75

for (i in seq_along(scatterplots)) {
  final_fig <- final_fig +
    annotation_custom(
      scatterplots[[i]],
      xmin = scatterplot_positions[[i]][1] - scatterplot_scale_x,
      xmax = scatterplot_positions[[i]][1] + scatterplot_scale_x,
      ymin = scatterplot_positions[[i]][2] - scatterplot_scale_y,
      ymax = scatterplot_positions[[i]][2] + scatterplot_scale_y
    )
}

final_fig

# -----------------------------
# 9) Export (force no clipping and use full width)
# -----------------------------
ggsave(
  "figures/FIGURE_2b_azores_inset_DCA.pdf",
  plot = final_fig,
  dpi = 300, width = 180, height = 90, units = "mm",
  limitsize = FALSE
)
```

### FIGURE 3 - Trophic structure classification
```{r fig.asp=2, fig.width=12}
# --------------------------------------------
# Global theme parameters (adjust once here)
# --------------------------------------------
axis_title_size <- 15
axis_text_size <- 13
plot_title_size <- 15

# --------------------------------------------
# Prepare Panel A: Ordination plot (removes legend for cleaner layout)
# --------------------------------------------
ordinations_fclusts_1 <- ordinations_fclusts_1 + 
  guides(fill = "none") +
  ggtitle("(a) Functional Structure Space") +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "pt"),
    axis.text = element_text(size = axis_text_size),
    axis.title = element_text(size = axis_title_size),
    plot.title = element_text(size = plot_title_size, face = "bold")
  )

# --------------------------------------------
# Prepare Panel B: Classification tree from PNG image
# --------------------------------------------
tree_grob <- grid::rasterGrob(png::readPNG("outputs/decision_tree_paleo.png"))
plot_decision_tree_paleo <- ggplot() +
  annotation_custom(tree_grob, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  theme_void() +
  ggtitle("(b) Classification Tree") +
  theme(plot.title = element_text(size = plot_title_size, face = "bold"))

# --------------------------------------------
# Prepare Panel C: Boxplots of trophic guilds across community structures
# --------------------------------------------
facet_colors <- c(
  "CTS1" = viridis::viridis(5, direction = -1)[1],
  "CTS2" = viridis::viridis(5, direction = -1)[2],
  "CTS3" = viridis::viridis(5, direction = -1)[3],
  "CTS4" = viridis::viridis(5, direction = -1)[4],
  "CTS5" = viridis::viridis(5, direction = -1)[5]
)

boxplots <- 
  bind_cols(global_matrix_merged_paleo, select(norm_abund, c(
    euplanktonic, high_profile, low_profile, motile, algivore, 
    detritivore, plantivore, predator))) %>%
  pivot_longer(
    cols = c(euplanktonic, high_profile, low_profile, motile, algivore, 
             detritivore, plantivore, predator),
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    amd_clusts = as.factor(amd_clusts),
    name = fct_relevel(name,
      "euplanktonic", "high_profile", "low_profile", "motile",
      "algivore", "detritivore", "plantivore", "predator"
    )
  ) %>%
  ggplot(aes(name, value)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(
    "euplanktonic" = "EU", "high_profile" = "HP", "low_profile" = "LP", "motile" = "MOT",
    "algivore" = "AL", "detritivore" = "DET", "plantivore" = "PL", "predator" = "PR"
  )) +
  theme_bw() +
  labs(x = "Community Trophic Guilds", y = "Relative Abundance") +
  guides(fill = "none") +
  facet_wrap2(
    ncol = 6,
    ~ factor(amd_clusts, labels = c("CTS1", "CTS2", "CTS3", "CTS4", "CTS5")),
    strip = strip_themed(
      text_x = element_text(color = "black"),
      background_x = list(
        "CTS1" = element_rect(fill = scales::alpha(facet_colors["CTS1"], 0.8)),
        "CTS2" = element_rect(fill = scales::alpha(facet_colors["CTS2"], 0.8)),
        "CTS3" = element_rect(fill = scales::alpha(facet_colors["CTS3"], 0.8)),
        "CTS4" = element_rect(fill = scales::alpha(facet_colors["CTS4"], 0.8)),
        "CTS5" = element_rect(fill = scales::alpha(facet_colors["CTS5"], 0.8))
      )
    )
  ) +
  ggtitle("(c) Guild Composition by Structure") +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "pt"),
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.text = element_text(size = axis_text_size),
    axis.title = element_text(size = axis_title_size),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2)
  )

# optional: if you want labels under the two groups
# library(ggtext)

guild_levels <- c(
  "euplanktonic", "high_profile", "low_profile", "motile",
  "GAP",
  "algivore", "detritivore", "plantivore", "predator"
)

guild_labels <- c(
  "euplanktonic" = "EU", "high_profile" = "HP", "low_profile" = "LP", "motile" = "MOT",
  "GAP" = "",
  "algivore" = "AL", "detritivore" = "DET", "plantivore" = "PL", "predator" = "PR"
)

boxplots <-
  bind_cols(global_matrix_merged_paleo,
            select(norm_abund, c(
              euplanktonic, high_profile, low_profile, motile,
              algivore, detritivore, plantivore, predator
            ))) %>%
  pivot_longer(
    cols = c(euplanktonic, high_profile, low_profile, motile,
             algivore, detritivore, plantivore, predator),
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    amd_clusts = as.factor(amd_clusts),
    name = factor(name, levels = setdiff(guild_levels, "GAP"))
  ) %>%
  # --- add a dummy row per facet to create a visual gap level ---
  bind_rows(
    distinct(., amd_clusts) %>%
      mutate(name = "GAP", value = NA_real_)
  ) %>%
  mutate(
    name = factor(as.character(name), levels = guild_levels)
  ) %>%
  ggplot(aes(name, value)) +
  # separator line between producers and consumers (after the 4th box)
  geom_vline(xintercept = 5, linewidth = 0.6, colour = "grey40") +
  geom_boxplot(na.rm = TRUE) +
  scale_x_discrete(labels = guild_labels, drop = FALSE) +
  theme_bw() +
  labs(x = "Community Trophic Guilds", y = "Relative Abundance") +
  guides(fill = "none") +
  facet_wrap2(
    ncol = 6,
    ~ factor(amd_clusts, labels = c("CTS1", "CTS2", "CTS3", "CTS4", "CTS5")),
    strip = strip_themed(
      text_x = element_text(color = "black"),
      background_x = list(
        "CTS1" = element_rect(fill = scales::alpha(facet_colors["CTS1"], 0.8)),
        "CTS2" = element_rect(fill = scales::alpha(facet_colors["CTS2"], 0.8)),
        "CTS3" = element_rect(fill = scales::alpha(facet_colors["CTS3"], 0.8)),
        "CTS4" = element_rect(fill = scales::alpha(facet_colors["CTS4"], 0.8)),
        "CTS5" = element_rect(fill = scales::alpha(facet_colors["CTS5"], 0.8))
      )
    )
  ) +
  ggtitle("(c) Guild Composition by Structure") +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "pt"),
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.text = element_text(size = axis_text_size),
    axis.title = element_text(size = axis_title_size),
    legend.background = element_rect(fill = "white", linewidth = 4, colour = "white"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.ticks = element_line(colour = "grey70", linewidth = 0.2)
  )

# --------------------------------------
# Prepare Panel D: Vegetation change with GAM smooth
# --------------------------------------
vegetation_smooth_plot <- ggplot(interpolated_df, aes(x = age_ce, y = estimate)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), fill = "steelblue", alpha = 0.2) +
  theme_minimal(base_size = 14) +
  labs(title = "Vegetation Change Regional Smooth", x = "Year CE", y = "Estimate")

df_p1_smooth <- ggplot_build(vegetation_smooth_plot)$data[[1]] %>%
  filter(!is.na(y)) %>%
  mutate(y_rescaled = scales::rescale(y, to = c(0, 1)))

df_p1_ribbon <- ggplot_build(vegetation_smooth_plot)$data[[2]] %>%
  filter(!is.na(ymin) & !is.na(ymax)) %>%
  mutate(
    ymin_rescaled = scales::rescale(ymin, to = c(0, 1)),
    ymax_rescaled = scales::rescale(ymax, to = c(0, 1))
  )

orig_range <- range(df_p1_smooth$y, na.rm = TRUE)

p_combined <- plot_regional_lake_balanced +
  geom_line(data = df_p1_smooth, aes(x = x, y = y_rescaled),
            inherit.aes = FALSE, color = "white", size = 2) +
  scale_y_continuous(
    name = "Proportion of CTS",
    sec.axis = sec_axis(
      trans = ~ scales::rescale(., from = c(0, 1), to = orig_range),
      name = "Vegetation Change (GAM smooth - Pollen signal)"
    )
  ) +
  theme_minimal(base_size = 12) +
  ggtitle("(d) Proportion of CTS Over Time") +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.text = element_text(size = axis_text_size),
    axis.title = element_text(size = axis_title_size)
  )

# --------------------------------------
# Prepare Panel E: Richness boxplots
# --------------------------------------
box_plot_clusts_mod <- box_plot_clusts_mod +
  scale_x_discrete(labels = c("CTS5", "CTS4", "CTS3", "CTS2", "CTS1")) +
  ggtitle("(e) Number of Species per CTS") +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.text = element_text(size = axis_text_size),
    axis.title = element_text(size = axis_title_size)
  )

# --------------------------------------------
# Define patchwork layout
# --------------------------------------------
layout <- "
AABB
AABB
CCCC
CCCC
"

# --------------------------------------------
# Assemble Panels A–C
# --------------------------------------------
figure_1_ord_rf_time_clusts <- (
  ordinations_fclusts_1 +
  plot_decision_tree_paleo +
  boxplots
) +
  patchwork::plot_layout(design = layout, heights = c(40, 40, 30), guides = "collect") +
  patchwork::plot_annotation() &
  theme(legend.position = 'none')

# --------------------------------------------
# Final assembly of all five panels
# --------------------------------------------
final_figure <- (
  figure_1_ord_rf_time_clusts /
  (
    (p_combined + theme(legend.position = "none")) +
    (box_plot_clusts_mod + theme(legend.position = "none"))
  ) +
  plot_layout(guides = "collect", heights = c(8, 5)) +
  plot_annotation(tag_levels = NULL)
)
print(final_figure)
# --------------------------------------------
# Save to file
# --------------------------------------------
ggsave(plot = final_figure, 
       filename = "figures/FIGURE_3_composite_figure.png", 
       dpi = 600,
       width = 12,
       height = 16)

ggsave(plot = final_figure, 
       filename = "figures/FIGURE_3_composite_figure.pdf", 
       dpi = 600,
       width = 12,
       height = 16)

```



### FIGURE 4 - Diversity Dynamics by Functional Group (Standardized Species Richness)
```{r fig.asp = 0.5, fig.width = 10}
# --------------------------------------------
# Common theme helpers for legends (prevents clipping)
# --------------------------------------------
legend_theme <- theme(
  legend.position = "bottom",
  legend.box = "vertical",                 # <-- stack the two legends
  legend.direction = "horizontal",
  legend.title = element_text(size = 11),
  legend.text  = element_text(size = 11),
  legend.key.width  = unit(1.3, "lines"),
  legend.key.height = unit(0.9, "lines"),
  legend.spacing.y  = unit(0.2, "lines"),
  plot.margin = ggplot2::margin(5.5, 5.5, 20, 5.5)  # <-- extra bottom space for legends
)

cons_labels <- c(
  "predator"      = "Predators",
  "detritivore"   = "Detritivores",
  "algivore"      = "Algivores",
  "plantivore"    = "Plantivores"
)

prod_labels <- c(
  "high_profile"  = "High-profile",
  "low_profile"   = "Low-profile",
  "motile"        = "Motile",
  "euplanktonic"  = "Euplanktonic"
)

# --------------------------------------------
# Consumer plot
# --------------------------------------------
cons_glob_div_plt <- cons_glob_div_plt_data %>%
  ggplot(aes(age_ce, standardized_species_richness)) +
  geom_smooth(aes(color = fgroup), se = TRUE) +
  geom_smooth(color = "red", linetype = "dotted", alpha = 0.7, se = FALSE) +
  scale_color_scico_d(
    name = "Consumers",
    palette = "lajolla",
    direction = -1,
    labels = cons_labels
  ) +
  labs(x = "Age (CE)", y = "Standardized Species Richness (0–1)") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = axis_title_size),
    axis.text  = element_text(size = axis_text_size),
    plot.title = element_text(size = plot_title_size, face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE)) +  # adjust nrow as you like
  lims(x = c(0, 2010))

# --------------------------------------------
# Producer plot
# --------------------------------------------
prod_glob_div_plt <- prod_glob_div_plt_data %>%
  ggplot(aes(age_ce, standardized_species_richness)) +
  geom_smooth(aes(color = fgroup), se = TRUE) +
  geom_smooth(color = "red", linetype = "dotted", alpha = 0.7, se = FALSE) +
  scale_color_viridis_d(
    name = "Producers",
    option = "D",
    direction = 1,
    alpha = 0.5,
    labels = prod_labels
  ) +
  labs(x = "Age (CE)", y = "Standardized Species Richness (0–1)") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = axis_title_size),
    axis.text  = element_text(size = axis_text_size),
    plot.title = element_text(size = plot_title_size, face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE)) +  # adjust nrow as you like
  lims(x = c(0, 2010))

# --------------------------------------------
# Combine (KEEP vertical), legends stacked at bottom
# IMPORTANT: do NOT use guides="collect" because scales differ
# --------------------------------------------
plot_div_per_fgroup_regional_scale <-
  (prod_glob_div_plt / cons_glob_div_plt) +
  plot_annotation(tag_levels = "a") &
  legend_theme

print(plot_div_per_fgroup_regional_scale)

# --------------------------------------------
# Save output (give enough width so legend text doesn’t clip)
# --------------------------------------------
ggsave(
  plot = plot_div_per_fgroup_regional_scale,
  filename = "figures/FIGURE_4_plot_div_per_fgroup_regional_scale.png",
  dpi = 600,
  width = 7.2, height = 11.5, units = "in"
)

ggsave(
  plot = plot_div_per_fgroup_regional_scale,
  filename = "figures/FIGURE_4_plot_div_per_fgroup_regional_scale.pdf",
  dpi = 600,
  width = 7.2, height = 11.5, units = "in"
)

```

## FIGURE 5 - Historical and Moving Window Variance Partitioning
```{r fig.width = 10, fig.asp = 1}
# Single-column layout: legends below each panel, consistent font scaling.
# --------------------------------------------
# Global figure controls (edit once)
# --------------------------------------------
base_size       <- 10   # overall text size (single-column friendly)
title_size      <- 11
axis_title_size <- 10
axis_text_size  <- 9
legend_title_sz <- 9
legend_text_sz  <- 8

# Single-column width (PNAS is typically ~3.3–3.5 in for 1-col figures)
# Use 3.5 if you want strict 1-col; 4.5 is safer for readability.
fig_width  <- 4.5
fig_height <- 10

# STEP 1: Define historical transition years
phase_lines <- c(750, 1050, 1450, 1750)

# --------------------------------------------
# A shared theme applied to ALL three panels
# --------------------------------------------
panel_theme <- theme_minimal(base_size = base_size) +
  theme(
    plot.title        = element_text(size = title_size, face = "bold", hjust = 0),
    axis.title        = element_text(size = axis_title_size),
    axis.text         = element_text(size = axis_text_size),
    legend.position   = "bottom",
    legend.title      = element_text(size = legend_title_sz),
    legend.text       = element_text(size = legend_text_sz),
    legend.key.height = unit(3.0, "mm"),
    legend.key.width  = unit(6.0, "mm"),
    plot.margin       = ggplot2::margin(3, 3, 3, 3)
  )

# --------------------------------------------
# Panel-specific legend helper
# (3 rows for a and b; and also for c)
# --------------------------------------------
legend_3rows <- guides(
  fill  = guide_legend(nrow = 3, byrow = TRUE),
  color = guide_legend(nrow = 3, byrow = TRUE),
  alpha = "none"
)

# --------------------------------------------
# Panel A
# --------------------------------------------
p_a <- historical_phase_varpart_sig +
  ggtitle("(a) Historical phases") +
  scale_x_discrete(labels = c("Phase 1", "Phase 2", "Phase 3", "Phase 4", "Phase 5")) +
  panel_theme +
  legend_3rows

# --------------------------------------------
# Panel B
# --------------------------------------------
p_b <- moving_window_varpart_sig +
  ggtitle("(b) Moving window") +
  panel_theme +
  legend_3rows

# --------------------------------------------
# Panel C
# --------------------------------------------
p_c <- effect_diff_plot_sig +
  ggtitle("(c) Climate vs VegChange") +
  panel_theme +
  guides(
    fill  = guide_legend(nrow = 3, byrow = TRUE),
    alpha = "none"
  )

# If any panel still shows an "NA" legend entry, uncomment:
# p_c <- p_c + scale_fill_discrete(na.translate = FALSE)

# --------------------------------------------
# Combine (single column) — keep legends under each panel
# --------------------------------------------
combined_varplot <- p_a / p_b / p_c +
  plot_layout(guides = "keep", heights = c(1, 1.2, 1))

print(combined_varplot)

# --------------------------------------------
# Save output (single-column)
# --------------------------------------------
ggsave(
  plot     = combined_varplot,
  filename = "figures/FIGURE_5_varparts_singlecol.png",
  dpi      = 600,
  width    = fig_width,
  height   = fig_height,
  units    = "in",
  bg       = "white"
)

ggsave(
  plot     = combined_varplot,
  filename = "figures/FIGURE_5_varparts_singlecol.pdf",
  dpi      = 600,
  width    = fig_width,
  height   = fig_height,
  units    = "in",
  bg       = "white"
)

```

# SUPPLEMENTARY FIGURES
## SUPPLEMENTARY FIGURE S2 - Climatic reconstructions
```{r fig.width=12, fig.asp=0.5}
nhst_df <- nhst_buntgen %>%
  arrange(age_ce) %>%
  mutate(
    sign = ifelse(Rmean >= 0, "positive", "negative")
  )

nao_df <- nao_hernandez %>%
  arrange(age_ce) %>%
  mutate(
    sign = ifelse(NAO_Median_Value >= 0, "positive", "negative")
  )

# --------------------------------------
# Plot 1: NH Summer Temperature (Büntgen tree-ring reconstruction)
# --------------------------------------
nhst_buntgen_plot <- ggplot(nhst_df, aes(x = age_ce, y = Rmean)) +
  geom_line(size = 1, color = "grey") +  # Main raw line
  geom_point(aes(color = sign), size = 1) +  # Positive (red) / negative (blue) deviations
  scale_color_manual(values = c("positive" = "red", "negative" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +  # zero reference

  # Add 50-year smoothed trend line
  geom_line(data = nhst_30, aes(x = age_ce, y = Rmean),
            color = "steelblue", linewidth = 2) +

  theme_minimal(base_size = 14) +
  labs(
    title = "NH Summer Temperature reconstruction (Tree Rings)",
    x = "Year CE",
    y = "Temperature Anomaly",
    color = NULL
  ) +
  theme(legend.position = "none")

# --------------------------------------
# Plot 2: North Atlantic Oscillation (NAO) median values
# --------------------------------------
nao_plot <- ggplot(nao_df, aes(x = age_ce, y = NAO_Median_Value)) +
  geom_line(size = 1, color = "grey") +  # Raw NAO data
  geom_point(aes(color = sign), size = 1) +  # Red/blue anomalies
  scale_color_manual(values = c("positive" = "red", "negative" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +

  # Add 50-year smoothed trend line
 #geom_line(data = nao_30, aes(x = age_ce, y = NAO_Median_Value),
 #          color = "steelblue", linewidth = 2) +

  theme_minimal(base_size = 14) +
  labs(
    title = "NAO Anomaly",
    x = "Year CE",
    y = "NAO Median",
    color = NULL
  ) +
  theme(legend.position = "none")

# --------------------------------------
# Custom Theme for Plot Tag Styling
# --------------------------------------
tag_theme <- theme(
  plot.tag.position = c(0, 1),
  plot.tag = element_text(
    hjust = 0,
    vjust = -2,
    face = "bold",
    size = 11,
    margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 10, unit = "pt")  # Left indent for tag
  ),
  plot.margin = ggplot2::margin(20, 10, 10, 10, unit = "pt")  # Standard margins
)

# --------------------------------------
# Assemble Tagged and Themed Panels
# --------------------------------------
nhst_buntgen_panel <- nhst_buntgen_plot + 
  labs(tag = "(a) NH Summer Temperature Anomaly reconstruction (Tree Rings)", title = NULL) + 
  tag_theme +
  theme(axis.title.x = element_blank())  # remove x-axis label

nao_panel <- nao_plot + 
  labs(tag = "(b) NAO Index", title = NULL) + 
  tag_theme +
  theme(axis.title.x = element_blank())  # remove x-axis label

# --------------------------------------
# Combine Panels Vertically with patchwork
# --------------------------------------
final_plot_2 <- 
  (nhst_buntgen_panel + nao_panel) +
  patchwork::plot_layout(ncol = 1, nrow = 3,
                         heights = c(20, 20, 20)) +  # Equal height for each plot
  patchwork::plot_annotation(
    tag_levels = NULL,
    theme = theme(
      plot.margin = ggplot2::margin(10, 10, 30, 10, unit = "pt"),  # extra bottom space
      plot.caption = element_text(hjust = 0.5, face = "bold", size = 12)
    )
  )

# --------------------------------------
# Export to PNG
# --------------------------------------

# Set PNG output for saving the tree plot
png("supplementary/supplementary_figure_S2.png", width = 8, height = 8, units = "in", res = 600)

# Plot
final_plot_2

# Close the graphics device
dev.off()
```

## SUPPLEMENTARY FIGURE S3 - Pollen GAM
```{r fig.width = 8, fig.asp = 1.5}

# --------------------------------------
# Plot 1: Group-Specific Pollen Trends (Tree vs Herb)
# --------------------------------------
group_trend_plot <- ggplot(combined_df_pollen, aes(x = age_ce, y = percent, color = group)) +
  geom_point(alpha = 0.2, size = 1) +  # Raw points with transparency
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 30),  # GAM smoothing with 30 knots
    se = TRUE,
    linewidth = 1.5
  ) +
  scale_color_manual(
    values = c("Tree" = "#3a6520", "Herb" = "#dbbc3f"),  # Custom color for each pollen group
    labels = c("Tree" = "Tree", "Herb" = "Herb"),
    name = "Pollen Group"
  ) +
  labs(
    title = "Vegetation change",
    x = "Age (CE)",
    y = "% Abundance"
  ) +
  coord_cartesian(xlim = c(0, 2010)) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold")
  )

# --------------------------------------
# Plot 2: Regional Pollen Turnover (GI model smooth)
# --------------------------------------
regional_smooth_plot <- interpolated_df %>%
  ggplot() +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, x = age_ce),
              alpha = 0.2) +  # Confidence interval
  geom_line(aes(x = age_ce, y = estimate), lwd = 1) +  # Smooth estimate line
  labs(y = "DCA1", title = "s(age_ce) GI model") +
  scale_color_manual("Time-Interval", values = lakes_col) +  # Optional (no visible use)
  theme_minimal(base_size = 12) +
  lims(y = c(-1, 1)) +
  labs(title = NULL) +
  theme(
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )

# --------------------------------------
# Panel A: Regional pollen turnover with label
# --------------------------------------
regional_panel_A <- regional_smooth_plot +
  labs(title = "(a) Regional Pollen Community Turnover") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  )

# --------------------------------------
# Panel B: Tree vs Herb pollen dynamics with label
# --------------------------------------
group_panel_B <- group_trend_plot +
  labs(title = "(b) Tree and Herb pollen dynamics") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  )

# --------------------------------------
# Combine Panels into Final Supplementary Figure
# --------------------------------------
final_combined_plot_pollen <- 
  regional_panel_A + 
  group_panel_B +
  patchwork::plot_layout(nrow = 2, heights = c(1, 1)) +
  patchwork::plot_annotation(
    theme = theme(
      plot.margin = ggplot2::margin(10, 10, 10, 10),
      plot.title = element_text(face = "bold", size = 14)
    )
  )

# --------------------------------------
# Export to PNG
# --------------------------------------
png("supplementary/supplementary_figure_S3_vegetation_change.png", 
    width = 7.29, height = 4.51, units = "in", res = 600)

# Render the plot to file
final_combined_plot_pollen

# Close the graphics device
dev.off()

```

## SUPPLEMENTARY FIGURE S4 - Pollen GAM by island
```{r fig.width = 8, fig.asp = 1.5}

# --------------------------------------
# Export to PNG
# --------------------------------------
png("supplementary/supplementary_figure_S4_pollen_by_island.png", 
    width = 7.29, height = 4.51, units = "in", res = 600)

# Render the plot to file
p_presence_rows_lakes

# Close the graphics device
dev.off()

```


### SUPPLEMENTARY FIGURE S5 - Community Trophic Structures over time
```{r fig.asp = 1.5, fig.width = 12}
# -----------------------------------------------
# Convert PNG tree plot into a ggplot-compatible object
# -----------------------------------------------
tree_grob_paleo <- grid::rasterGrob(png::readPNG("outputs/decision_tree_paleo.png"))

# Create a ggplot with the decision tree image embedded as a background
tree_image_panel <- ggplot() +
  annotation_custom(tree_grob_paleo, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  theme_void() +
  ggtitle("Functional group classification by ecosystem type")  # Title for Panel D

# -----------------------------------------------
# Clean up lake-scale density plot (remove redundant legend)
# -----------------------------------------------
lake_scale_density_plot <- plot_density_fclusts_time_lines_1B + 
  guides(fill = "none")  # Remove fill legend (i.e., for functional groups)

# -----------------------------------------------
# Define layout structure for patchwork combination
# -----------------------------------------------
patchwork_layout_design <- "
AAAA
DDDD
DDDD
DDDD
"

# -----------------------------------------------
# Assemble Supplementary Figure 4
# -----------------------------------------------

# Panel A: Regional scale
panel_regional_stacked <- plot_stacked_fclusts_time_lines +
  ggtitle(" (a) Community Trophic Structures over time [regional scale]") +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "pt"),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Panel B: Lake scale
panel_lake_density <- lake_scale_density_plot +
  ggtitle(" (b) Community Trophic Structures over time [lake scale]") +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "pt"),
    legend.position = "top",
    legend.key.size = unit(0.5, "lines"),
    legend.text = element_text(size = 10),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
# Combine panels using patchwork with layout control
supplementary_figure_5 <- 
  (panel_regional_stacked + panel_lake_density) +
  patchwork::plot_layout(
    design = patchwork_layout_design,
    heights = c(40, 80),
    guides = "collect"
  ) +
  #patchwork::plot_annotation(tag_levels = 'A') &
  theme(
    plot.margin = unit(c(0, 0, 0, 0), "pt"),  # Minimize outer spacing
    legend.position = 'bottom'               # Merge legends at the bottom
  )

# -----------------------------------------------
# Display the combined supplementary figure
# -----------------------------------------------
print(supplementary_figure_5)

# -----------------------------------------------
# Save to file
# -----------------------------------------------
ggsave(
  plot = supplementary_figure_5,
  filename = "supplementary/supplementary_figure_S5.png",
  dpi = 600,
  width = 12,
  height = 16
)

```

### SUPPLEMENTARY FIGURE S6 - Variable Importance Plot
```{r}

# Set PNG output for saving the tree plot
png("supplementary/supplementary_figure_S6_varimpplot.png", width = 7.29, height = 4.51, units = "in", res = 600)

# Plot variable importance from random forest
varImpPlot(RFfit_paleo, main = "")

# Close the graphics device
dev.off()

```

### SUPPLEMENTARY FIGURE S7 - Global RDA
```{r fig.width = 12, fig.asp = 1}

# Set PNG output for saving the tree plot
png("supplementary/supplementary_figure_S7_global_rda.png", width = 8, height = 8, units = "in", res = 600)

# Plot Global Rda with selected variables
global_rda

# Close the graphics device
dev.off()

```

### SUPPLEMENTARY FIGURE S9 - Null model
```{r fig.width = 12, fig.asp = 1}

# Set PNG output for saving the tree plot
png("supplementary/supplementary_figure_S9_null_model.png", width = 8, height = 8, units = "in", res = 600)

# Plot Global Rda with selected variables
combined_plot

# Close the graphics device
dev.off()

```


